---
title: "Performance assessment of survival prediction models - extended code"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 5
  keep_text: true
  pandoc_args: --webtex
---
```{r setup, include=FALSE}
# Knitr options
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.path = "imgs/03_predsurv_extended/"
)
```

## Goals
In this document, we assume that individual data of the development and validation set are both available.

In particular:   
1. To develop a risk prediction model with a time-to-event outcome;  
2. To assess the prediction performance of a model with a time-to-event outcome;  
3. To assess the potential clinical utility of a risk prediction model with time-to-event outcome.

The extended code basically evaluates the prediction performance at a fixed time horizon (e.g. 5 years) and also for the entire follow-up time.

### Load packages and import data
We following libraries are needed to achieve the following goals, if you have not them installed, please use
install.packages('') (e.g. install.packages('survival')) or use the user-friendly approach if you are using RStudio.

```{r, wdlib, message=FALSE,warning=FALSE}

# Use pacman to check whether packages are installed, if not load
if (!require("pacman")) install.packages("pacman")
library(pacman)

pacman::p_load(
  rio,
  survival,
  rms,
  mstate,
  sqldf,
  pec,
  riskRegression,
  survAUC,
  survivalROC,
  timeROC,
  plotrix,
  splines,
  knitr,
  table1,
  kableExtra,
  gtsummary,
  boot,
  tidyverse,
  rsample,
  gridExtra,
  webshot
)
```

The two primary datasets from the Rotterdam study and the
German Breast Cancer Study Group (GBSG) are 
included as example data sets in the survival package.
The Rotterdam data has separate variables for time to recurrence
of disease and time to death, and these are combined into a single
endpoint of recurrence free survival (RFS), while the gbsg data
set has RFS as its single endpoint.
The follow-up time is converted from days to years, that being a more
natural unit for any plots; it does not change any fits.
A categorical variable for nodes is also added.


### Descriptive statistics
```{r, import,fig.align='center'}

# Development data
# Recurrence free survival (RFS) is the time until the earlier of
#  recurrence or death
rotterdam$ryear <- rotterdam$rtime/365.25  # time in years
rotterdam$rfs <- with(rotterdam, pmax(recur, death))

# variables used in the analysis
pgr99 <- quantile(rotterdam$pgr, .99) # there is a large outlier of 5000
rotterdam$pgr2 <- pmin(rotterdam$pgr, pgr99) # Winsorized value
rotterdam$csize <- rotterdam$size           # categorized size
rotterdam$cnode <- cut(rotterdam$nodes, c(-1,0, 3, 51),
                       c("0", "1-3", ">3"))   # categorized node

# Save in the data the restricted cubic spline term using Hmisc::rcspline.eval() package
rcs3_pgr <- rcspline.eval(rotterdam$pgr2, knots = c(0, 41, 486))
attr(rcs3_pgr, "dim") <- NULL
attr(rcs3_pgr, "knots") <- NULL
rotterdam$pgr3 <- rcs3_pgr

# Validation data
gbsg$ryear <- gbsg$rfstime/365.25
gbsg$rfs   <- gbsg$status           # the GBSG data contains RFS
gbsg$cnode <- cut(gbsg$nodes, c(-1, 0, 3, 51),
                       c("0", "1-3", ">3"))   # categorized node
gbsg$csize <- cut(gbsg$size,  c(-1, 20, 50, 500), #categorized size
                  c("<=20", "20-50", ">50"))
gbsg$pgr2 <- pmin(gbsg$pgr, pgr99) # Winsorized value


# Restricted cubic spline for PGR
rcs3_pgr <- rcspline.eval(gbsg$pgr2, knots = c(0, 41, 486))
attr(rcs3_pgr, "dim") <- NULL
attr(rcs3_pgr, "knots") <- NULL
gbsg$pgr3 <- rcs3_pgr


# Much of the analysis will focus on the first 5 years: create
#  data sets that are censored at 5
temp <- survSplit(Surv(ryear, rfs) ~ ., data = rotterdam, cut = 5,
                  episode = "epoch")
rott5 <- subset(temp, epoch == 1)  # only the first 5 years
temp <- survSplit(Surv(ryear, rfs) ~ ., data = gbsg, cut = 5,
                  episode = "epoch")
gbsg5 <- subset(temp, epoch == 1)

# Relevel
rott5$cnode <- relevel(rotterdam$cnode, "1-3")
gbsg5$cnode <- relevel(gbsg$cnode, "1-3")
```


```{r tab1_bis, echo=FALSE}
rsel <- rotterdam[, c("pid", "csize", "cnode", "grade", "age", "pgr")]
vsel <- gbsg[, c("pid", "csize", "cnode", "grade", "age", "pgr")]
rsel$dt <- 1
vsel$dt <- 2
cdata <- rbind(rsel, vsel)
cdata$dt <- factor(cdata$dt,
  levels = c(1, 2),
  labels = c("Development dataset", "Validation dataset")
)

label(cdata$csize) <- "Size"
label(cdata$cnode) <- "Number of nodes"
label(cdata$grade) <- "Grade of tumor"
label(cdata$age) <- "Age"
label(cdata$pgr) <- "PGR"
label(cdata$dt) <- "Dataset"

units(cdata$csize) <- "mm"
units(cdata$age) <- "years"
units(cdata$pgr) <- "ng/mL"


gtsummary::tbl_summary(
  data = cdata %>% select(-pid),
  label = list(age ~ "Age (years)", csize ~ "Size (cm)", 
               pgr ~ "PGR (ng/mL)",
               grade ~ "Grade"),
  by = "dt", 
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
) %>% 
  gtsummary::as_kable_extra() %>% 
  kableExtra::kable_styling("striped") 
```


## Goal 1 - Develop a risk prediction model with a time-to-event outcome
Prediction models are useful to provide the estimated probability of a specific outcome using personal information.
In many studies, especially in medicine, the main outcome under assessment is the time to an event of interest defined generally as survival time. Prognostic models for survival end points, such as recurrence or progression of disease, need to account for drop out during follow-up. Patients who have not experienced the event of interest are censored observations. Cox regression analysis is the most popular statistical model to deal with such data in oncology and other medical research.  

### 1.1 Preliminary investigation - survival and censoring curves in the development and validation data
First, we draw the survival and the censoring curves of the development and validation data
```{r, surv, fig.align='center'}
# Development set
sfit_rott <- survfit(Surv(ryear, rfs == 1) ~ 1, 
                     data = rotterdam) # survival
sfit_rott_c <- survfit(Surv(ryear, rfs == 0) ~ 1, 
                       data = rotterdam) # censoring

par(xaxs = "i", yaxs = "i", las = 1)
plot(sfit_rott, conf.int = FALSE, lwd = 2, xlab = "Years", bty = "n")
lines(sfit_rott_c, conf.int = FALSE, col = 2, lwd = 2)
legend(11, .9, c("Death", "Censoring"), col = 1:2, lwd = 2, bty = "n")
title("Development set")


# Validation set
sfit_gbsg <- survfit(Surv(ryear, rfs == 1) ~ 1, 
                     data = gbsg) # survival
sfit_gbsg_c <- survfit(Surv(ryear, rfs == 0) ~ 1, 
                       data = gbsg) # censoring

par(xaxs = "i", yaxs = "i", las = 1)
plot(sfit_gbsg, conf.int = FALSE, 
     lwd = 2, xlab = "Years", bty = "n", xlim = c(0, 8))
lines(sfit_gbsg_c, conf.int = FALSE, 
      col = 2, lwd = 2)
legend("bottomleft", c("Death", "Censoring"), col = 1:2, lwd = 2, bty = "n")
title("Validation set")
```
A number of 2982 patients were included to develop the risk prediction model for survival with a median follow-up of 9 years. The 5-year survival was 56% (95% CI: 55-58%). 
A number of 686 patients were selected to externally validate the risk prediction model. The median survival in the validation data was 4.5 years. The median survival was 5 years while the 5-year survival was 49% (95% CI: 45-54%).


### 1.2 Secondary investigation - check non-linearity of continuous predictors
The potential non-linear relation between continuous predictors (i.e. progesterone level) and the outcome should be investigated before developing a risk prediction model. Non-linearity of continuous predictors can be checked using splines.  
Physically, a spline is a flexible wood or metal strip, which is passed through a set of fixed points (knots) in order to approximate a curve. The most common computational approximation to this is a cubic smoothing spline which is cubic between the knot points, and constrained to be linear beyond the two end knots. 
For the restricted cubic spline using rms::rcs() R package::function(), the position of the knots are defined at 10<sup>th</sup>,50<sup>th</sup> and 90<sup>th</sup> quantile of the continuous predictor distribution. For more details see Frank Harrell's book 'Regression Model Strategies' on page 27 (second edition).  
The user can specify the positions of the knots instead of using the default calculation of the knots proposed in the book of Frank Harrell.
To deal with very large influential value, we winzorize progesterone level to the 90<sup>th</sup>  percentile.

```{r,ff, warning=FALSE, fig.align='center'}
dd <- datadist(rotterdam)
options(datadist = "dd")
fit_pgr <- cph(Surv(ryear, rfs) ~ rcs(pgr2), 
               data = rotterdam, x = T, y = T, surv = T)
plot(Predict(fit_pgr))
options(datadist = NULL)
```

We should model the progesterone level using a three-knot restricted cubic spline. We save the spline in the development and validation data.


### 1.3 Model development - first check - the proportional hazard (PH) assumption
We now examine the fits in a more careful way by checking the proportionality of the hazards of the Cox regression model.
Firstly, we fit the first prediction model in the development data using size, node, grade. Then, we check the PH assumption.

```{r,ph, message=FALSE, warning=FALSE,fig.align='center'}
fit1_ph <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade, 
                data = rotterdam, x = T, y = T)


zp1 <- cox.zph(fit1_ph, transform = "identity")
kable(round(zp1$table, 3)) %>% kable_styling("striped", position = "center")

oldpar <- par(mfrow = c(2, 2), mar = c(5, 5, 1, 1))
for (i in 1:3) {
  plot(zp1[i], resid = F)
  abline(0, 0, lty = 3)
}
par(oldpar)
```

The statistical tests show strong evidence of non-proportionality. Since the number of death is large the formal tests are quite sensitive, however, and it is important to also examine the graphs.  
These show an estimated coefficient as a function of time. As a further follow-up we will divide the data into 3 epochs of 0-5, 5-10, and 10+ years, fitting a separate model to each.

```{r, epoch,warning=FALSE}
# Development
edata <- survSplit(Surv(ryear, rfs) ~ .,
  data = rotterdam, cut = c(5, 10),
  episode = "epoch"
)
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = edata[edata$epoch == 1, ], x = T, y = T
)
efit2 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = edata[edata$epoch == 2, ], x = T, y = T
)
efit3 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = edata[edata$epoch == 3, ], x = T, y = T
)
```

```{r,ph_table, echo=FALSE}
res_efit <- round(rbind(e1 = coef(efit1), 
                        e2 = coef(efit2), 
                        e3 = coef(efit3)), 2)
rownames(res_efit) <- c("Epoch 1: 0-5 years", 
                        "Epoch 2: 5-10 years", 
                        "Epoch 3: >10 years")
kable(res_efit) %>% kable_styling("striped", position = "center")

options(prType = "html")
tt1 <- table(edata$epoch, edata$rfs, dnn = c("Epoch", "Status"))
rownames(tt1) <- c("Epoch 1: 0-5 years", 
                   "Epoch 2: 5-10 years", 
                   "Epoch 3: >10 years")
kable(tt1,
  col.names = c("Censored", "Event"),
  row.names = TRUE
) %>% kable_styling("striped", position = "center")
```

A drastic change in the size coefficients across all epochs is apparent,
along with a major reduction in the nodes coefficient in epoch 3. As an ameleoration of this we will refit the model using only the first epoch, which includes most of the recurrences and deaths.  
We applied the administrative censoring at 5 years in the development data and we assessed the prediction performance of the prognostic model at 5 years. The 5-year horizon was chosen because it is also a common prediction time horizon in clinical practice.  The hazards in the development data seem not totally proportional within 5 years but minor deviation of proportionality were considered acceptable. 

### 1.4 Model development - fit the risk prediction models
We develop the risk prediction model in the development data considering the first 5-year follow-up to minimize the violation of proportional hazard including size, nodel and grade. The second model also includes the progesterone level modelled using a 3-knot restricted cubic spline.  
We also administratively censored the validation data at 5 years.


```{r, model_development, fig.align='center'}
# Consider the first 5-year epoch in the development set
# Refit the model
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

```


Below the results of the models:  

+ Classical model:  
```{r, summary_m1, fig.align='center',warning=FALSE, echo = FALSE}
dd <- datadist(rott5)
options(datadist = "dd")
options(prType = "html")
fit_cph_rott5 <- cph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = rott5, x = T, y = T, surv = T
)
print(fit_cph_rott5)
```

+ Extended model:  

```{r, summary_m2, fig.align='center',warning=FALSE , echo = FALSE}
options(prType = "html")
fit_cph_rott5_pgr <- update(fit_cph_rott5, . ~ . + rcs(pgr2, c(0, 41, 486)))
print(fit_cph_rott5_pgr)
options(datadist = NULL)
```

The coefficients of the models indicated that higher size, higher number of positive lymph nodes and higher grade is more associate with poorer prognosis. The association of the progesterone biomarker and the outcome is non-linear as investigated previously.


## Goal 2 - Assessing performance in survival prediction models
The performance of a risk prediction models may be evaluated through:  
  
  + discrimination:  the ability of the model to identify patients with and without the outcome and it requires the coefficients (or the log of the hazard ratios) of the developed risk prediction model to be evaluated.   

+ calibration: the agreement between observed and predicted probabilities. It requires the baseline (cumulative) hazard or survival.     
+ overall performance measures: as a combination of discrimination and calibration and/or as a measure of the explained variation;  

+ overall measures: combination of discrimination and calibration measures

Unfortunately, a few publications report the complete baseline (cumulative) hazard or survival or even the baseline (cumulative) hazard or survival at fixed time horizon _t_.
If we had both individual data of the development and validation, a complete assessment of discrimination and calibration would be possible. We could evaluate the prediction performance of a risk prediction model at a fixed time horizon(s) _t_ and for the complete follow-up time.
In risk prediction, physicians typically focus on one or more clinically relevant time horizons to inform subjects about their risk. For this reasons, according to information available, different levels of validation assessment are possible.
Here we aim to assess the prediction performance of a risk prediction model with time-to-event outcome in case all individual data are available and in case of only the model equation of a fixed time horizon (i.e. at 5 years) is provided including the baseline survival.

### 2.1 Overall performance measures
Some overall performance measures are proposed using survival data:
    
+ Brier score: it is the squared differences between observed and predicted values at fixed time point (e.g. at 5 years);
  
+ Scaled Brier score (also known as Index of prediction accuracy - IPA): it improves interpretability by scaling the Brier Score with the Brier of the null model.  
  
  
First we save elements need to calculate the performance measures as the linear predictor and the predicted survival at 5 years in the development and validation data. Secondly, we create B bootstrap data to calculate percentile bootstrap confidence intervals needed for some performance measures.


```{r, overall, warning=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# NOTE: waiting for brier score function created by Terry.

# Development set (apparent Brier and IPA) without pgr
score_rott5 <-
  Score(list("cox_development" = efit1),
    formula = Surv(ryear, rfs) ~ 1, 
    data = rott5, 
    conf.int = TRUE, 
    times = 4.95,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )

# Validation set without pgr
score_gbsg5 <-
  Score(list("cox_validation" = efit1),
    formula = Surv(ryear, status) ~ 1, 
    data = gbsg5, 
    conf.int = TRUE,
    times = 4.95,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )

# Development set (apparent Brier and IPA) with pgr
score_rott5_pgr <-
  Score(list("cox_development" = efit1_pgr),
    formula = Surv(ryear, rfs) ~ 1, 
    data = rott5, 
    conf.int = TRUE, 
    times = 4.95,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )

# Validation set with pgr
score_gbsg5_pgr <-
  Score(list("cox_validation" = efit1_pgr),
    formula = Surv(ryear, rfs) ~ 1, 
    data = gbsg5, 
    conf.int = TRUE, 
    times = 4.95,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )



# Bootstrap for IPA -------------------------
B <- 10
set.seed
rott5_boot <- bootstraps(rott5, times = B)
gbsg5_boot <- bootstraps(gbsg5, times = B)

# IPA confidence intervals
# NOTE: computational time is too long to compute them when the number of bootstrap replications is high

# Bootstrap for IPA - basic model
score_boot <- function(split) {
  Score(list("model" = efit1),
    formula = Surv(ryear, rfs) ~ 1, 
    data = analysis(split), 
    conf.int = FALSE, 
    times = 4.95,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )$Brier$score[model == "model"][["IPA"]]
}

# Bootstrap for IPA - extended model with PGR
score_boot_pgr <- function(split) {
  Score(list("model" = efit1_pgr),
    formula = Surv(ryear, rfs) ~ 1, 
    data = analysis(split), 
    conf.int = FALSE, 
    times = 4.95,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )$Brier$score[model == "model"][["IPA"]]
}



rott5_boot <- rott5_boot %>% mutate(
  IPA = map_dbl(splits, score_boot),
  IPA_pgr = map_dbl(splits, score_boot_pgr)
) # Development dataset

gbsg5_boot <- gbsg5_boot %>% mutate(
  IPA = map_dbl(splits, score_boot),
  IPA_pgr = map_dbl(splits, score_boot_pgr)
) # Validation dataset

```

```{r, int_val, echo=FALSE, warning=FALSE}
source(here::here('Functions/internal_cv.R'))
# See legend of the argument in the Functions folder
```

```{r, int_val_calc, echo=FALSE, warning=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

int_val <- bootstrap_cv(db = rott5, 
                        B = 10,
                        time = "ryear",
                        status = "rfs",
                        formula_model = "Surv(ryear, rfs) ~ csize + 
                        cnode + 
                        grade",
                        formula_ipcw = "Surv(ryear, rfs) ~ 1",
                        pred.time = 4.95)

int_val_pgr <- bootstrap_cv(db = rott5, 
                        B = 10,
                        time = "ryear",
                        status = "rfs",
                        formula_model = "Surv(ryear, rfs) ~ csize + 
                        cnode + 
                        grade + pgr2 + pgr3",
                        formula_ipcw = "Surv(ryear, rfs) ~ 1",
                        pred.time = 4.95)
```

```{r, overall_table, echo=FALSE}
# Table overall measures
alpha <- .05
k <- 2 # number of digits
res_ov <- matrix(unlist(c(
  score_rott5$Brier$score[model == "cox_development"][["Brier"]],
  score_rott5$Brier$score[model == "cox_development"][["lower"]],
  score_rott5$Brier$score[model == "cox_development"][["upper"]],
  
  int_val["Brier corrected"],
  NA,
  NA,
  
  score_rott5_pgr$Brier$score[model == "cox_development"][["Brier"]],
  score_rott5_pgr$Brier$score[model == "cox_development"][["lower"]],
  score_rott5_pgr$Brier$score[model == "cox_development"][["upper"]],
  
  int_val_pgr["Brier corrected"],
  NA,
  NA,
  
  score_gbsg5$Brier$score[model == "cox_validation"][["Brier"]],
  score_gbsg5$Brier$score[model == "cox_validation"][["lower"]],
  score_gbsg5$Brier$score[model == "cox_validation"][["upper"]],
  
  
  score_gbsg5_pgr$Brier$score[model == "cox_validation"][["Brier"]],
  score_gbsg5_pgr$Brier$score[model == "cox_validation"][["lower"]],
  score_gbsg5_pgr$Brier$score[model == "cox_validation"][["upper"]],
  
  score_rott5$Brier$score[model == "cox_development"][["IPA"]],
  quantile(rott5_boot$IPA, probs = alpha / 2),
  quantile(rott5_boot$IPA, probs = 1 - alpha / 2),
  
  int_val["IPA corrected"],
  NA,
  NA,
  
  score_rott5_pgr$Brier$score[model == "cox_development"][["IPA"]],
  quantile(rott5_boot$IPA_pgr, probs = alpha / 2),
  quantile(rott5_boot$IPA_pgr, probs = 1 - alpha / 2),
  
  int_val_pgr["IPA corrected"],
  NA,
  NA,
  
  score_gbsg5$Brier$score[model == "cox_validation"][["IPA"]],
  quantile(gbsg5_boot$IPA, probs = alpha / 2),
  quantile(gbsg5_boot$IPA, probs = 1 - alpha / 2),
  
  score_gbsg5_pgr$Brier$score[model == "cox_validation"][["IPA"]],
  quantile(gbsg5_boot$IPA_pgr, probs = alpha / 2),
  quantile(gbsg5_boot$IPA_pgr, probs = 1 - alpha / 2)
)),
  nrow = 2, 
  ncol = 18, 
  byrow = T, 
  dimnames = list(
    c("Brier", "IPA"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"), 6)
)
)

res_ov <- round(res_ov, 2) # Digits
kable(res_ov) %>%
  kable_styling("striped", position = "center") %>%
  add_header_above(c(" " = 1, 
                     "Apparent" = 3, 
                     "Internal" = 3, 
                     "Apparent + PGR" = 3, 
                     "Internal + PGR" = 3, 
                     "External" = 3, 
                     "External + PGR" = 3))
```

As expected the overall performance measures were lower in the external validation. Including information about PGR slightly improved the overall performance.

### 2.2 Discrimination measures

NOTE: to be adjusted see the original paper
Discrimination is the ability to differentiate between subjects who have the outcome and subjects who do not.
In prognostic modelling, discrimination reflects separation between survival curves for individuals or groups.

Concordance can be assessed over several different time intervals:

* the entire range of the data
* a 5 year window corresponding to our target assessment point

Clearly the last of these is most relevant

This is easy to compute using the concordance function in the survival
package.
There is some uncertainty in the literature about the original Harrell
formulation verus Uno's suggestion to re-weight the time scale by the
factor $1/G^2(t)$ where $G$ is the censoring distribution.
There is more detailed information in the concordance vignette found in the
survival package.
  
We used the time horizon at 4.95 and not 5 years since controls are considered patients at risk after the time horizon and we administratively censored at 5 years to minimize the violation of PH assumption (see paragraph 1.3).  

The Uno's C-index provides a measure of an overall discrimination over the follow-up time. The Uno's time-dependent AUC measures discrimination at a specific time horizon *t* defining cases and controls of censored data at a fixed time horizon *t*.


More details are in the paper and in the references.

```{r, concordance, message=FALSE,warning=FALSE, fig.align='center',include=FALSE,echo=FALSE, eval=FALSE}

# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

## Save elements needed to estimate concordance indexes
# Development data 
rott5$lp <- predict(efit1, newdata = rott5)
rott5$lp_pgr <- predict(efit1_pgr, newdata = rott5)

# Validation data 
gbsg5$lp <- predict(efit1, newdata = gbsg5)
gbsg5$lp_pgr <- predict(efit1_pgr, newdata = gbsg5)

# Harrell C - development
harrell_C_rott5 <- concordance(Surv(ryear, rfs) ~ lp, 
                               rott5, 
                               reverse = TRUE)

harrell_C_rott5_pgr <- concordance(Surv(ryear, rfs) ~ lp_pgr, 
                               rott5, 
                               reverse = TRUE)

# Harrell C - validation
harrell_C_gbsg5 <- concordance(Surv(ryear, rfs) ~ lp, 
                               gbsg5, 
                               reverse = TRUE)

harrell_C_gbsg5_pgr <- concordance(Surv(ryear, rfs) ~ lp_pgr, 
                               gbsg5, 
                               reverse = TRUE)

# Uno's C - development
Uno_C_rott5 <- concordance(Surv(ryear, rfs) ~ lp, 
                           rott5, 
                           reverse = TRUE,
                           timewt = "n/G2")

Uno_C_rott5_pgr <- concordance(Surv(ryear, rfs) ~ lp_pgr, 
                           rott5, 
                           reverse = TRUE,
                           timewt = "n/G2")

# Uno's C - validation
Uno_C_gbsg5 <- concordance(Surv(ryear, rfs) ~ lp, 
                           gbsg5, 
                           reverse = TRUE,
                           timewt = "n/G2")

Uno_C_gbsg5_pgr <- concordance(Surv(ryear, rfs) ~ lp_pgr, 
                           gbsg5, 
                           reverse = TRUE,
                           timewt = "n/G2")

```



```{r, res_UnoC,echo=FALSE,include=FALSE,eval=FALSE}
alpha <- .05
k <- 2
res_C <- matrix(c(
  harrell_C_rott5$concordance,
  harrell_C_rott5$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5$var) ,
  harrell_C_rott5$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5$var),
  
  int_val["Harrell C corrected"],
  NA,
  NA,
  
  harrell_C_rott5_pgr$concordance,
  harrell_C_rott5_pgr$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5_pgr$var) ,
  harrell_C_rott5_pgr$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5_pgr$var),
  
  int_val_pgr["Harrell C corrected"],
  NA,
  NA,
  
  harrell_C_gbsg5$concordance,
  harrell_C_gbsg5$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_gbsg5$var) ,
  harrell_C_gbsg5$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_gbsg5$var),
  
  harrell_C_gbsg5_pgr$concordance,
  harrell_C_gbsg5_pgr$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_gbsg5_pgr$var) ,
  harrell_C_gbsg5_pgr$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_gbsg5_pgr$var),
  
  Uno_C_rott5$concordance,
  harrell_C_rott5$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5$var) ,
  harrell_C_rott5$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5$var),
  
  int_val["Harrell C corrected"],
  NA,
  NA,
  
  Uno_C_rott5_pgr$concordance,
  Uno_C_rott5_pgr$concordance - 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_rott5_pgr$var) ,
  Uno_C_rott5_pgr$concordance + 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_rott5_pgr$var),
  
  int_val_pgr["Uno C corrected"],
  NA,
  NA,
  
 
  Uno_C_gbsg5$concordance,
  Uno_C_gbsg5$concordance - 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_gbsg5$var) ,
  Uno_C_gbsg5$concordance + 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_gbsg5$var),
  
  Uno_C_gbsg5_pgr$concordance,
  Uno_C_gbsg5_pgr$concordance - 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_gbsg5_pgr$var) ,
  Uno_C_gbsg5_pgr$concordance + 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_gbsg5_pgr$var)
),
  
  nrow = 2,
  ncol = 18,
  byrow = T,
  
  dimnames = list(
    c("Harrell C", "Uno C"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"), 6)
  )
)

res_C <- round(res_C, 2) # Digits
kable(res_C) %>%
  kable_styling("striped", position = "center") %>%
  add_header_above(c(" " = 1, 
                     "Apparent" = 3, 
                     "Internal" = 3, 
                     "Apparent + PGR" = 3, 
                     "Internal + PGR" = 3, 
                     "External" = 3, 
                     "External + PGR" = 3))
```


```{r, AUC, message=FALSE, warning=FALSE}

# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Time-dependent AUC (in Table 3 called Uno's TD AUC at 5 years) ###
# Uno's time-dependent Area Under the Curve
# Development
Uno_rott5 <-
  timeROC(
    T = rott5$ryear, 
    delta = rott5$rfs,
    marker = predict(efit1, newdata = rott5),
    cause = 1, 
    weighting = "marginal", 
    times = 4.95,
    iid = TRUE
  )

Uno_rott5_pgr <-
  timeROC(
    T = rott5$ryear, 
    delta = rott5$rfs,
    marker = predict(efit1_pgr, newdata = rott5),
    cause = 1, 
    weighting = "marginal", 
    times = 4.95,
    iid = TRUE
  )

# Validation
Uno_gbsg5 <-
  timeROC(
    T = gbsg5$ryear, 
    delta = gbsg5$rfs,
    marker = predict(efit1, newdata = gbsg5),
    cause = 1, 
    weighting = "marginal", 
    times = 4.95,
    iid = TRUE
  )

Uno_gbsg5_pgr <-
  timeROC(
    T = gbsg5$ryear, 
    delta = gbsg5$rfs,
    marker = predict(efit1_pgr, newdata = gbsg5),
    cause = 1, 
    weighting = "marginal", 
    times = 4.95,
    iid = TRUE
  )
# NOTE: if you have a lot of data n > 2000, standard error computation may be really long.
# In that case, please use bootstrap percentile to calculate confidence intervals.
```

```{r, discr_table, echo=FALSE}
# Save results
alpha <- .05
k <- 2
res_AUC <- matrix(c(
 
  Uno_rott5$AUC["t=4.95"],
  Uno_rott5$AUC["t=4.95"] -
    qnorm(1 - alpha / 2) * Uno_rott5$inference$vect_sd_1["t=4.95"],
  Uno_rott5$AUC["t=4.95"] +
    qnorm(1 - alpha / 2) * Uno_rott5$inference$vect_sd_1["t=4.95"],
  
  int_val["AUC corrected"],
  NA,
  NA,
  
  Uno_rott5_pgr$AUC["t=4.95"],
  Uno_rott5_pgr$AUC["t=4.95"] -
    qnorm(1 - alpha / 2) * Uno_rott5_pgr$inference$vect_sd_1["t=4.95"],
  Uno_rott5_pgr$AUC["t=4.95"] +
    qnorm(1 - alpha / 2) * Uno_rott5_pgr$inference$vect_sd_1["t=4.95"],
  
  int_val_pgr["AUC corrected"],
  NA,
  NA,
  
  Uno_gbsg5$AUC["t=4.95"],
  Uno_gbsg5$AUC["t=4.95"] -
    qnorm(1 - alpha / 2) * Uno_gbsg5$inference$vect_sd_1["t=4.95"],
  Uno_gbsg5$AUC["t=4.95"] +
    qnorm(1 - alpha / 2) * Uno_gbsg5$inference$vect_sd_1["t=4.95"],
  
  Uno_gbsg5_pgr$AUC["t=4.95"],
  Uno_gbsg5_pgr$AUC["t=4.95"] -
    qnorm(1 - alpha / 2) * Uno_gbsg5_pgr$inference$vect_sd_1["t=4.95"],
  Uno_gbsg5_pgr$AUC["t=4.95"] +
    qnorm(1 - alpha / 2) * Uno_gbsg5_pgr$inference$vect_sd_1["t=4.95"]
),

  nrow = 1, 
  ncol = 18, 
  byrow = T,

  dimnames = list(
    c("Uno AUC"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"), 6)
)
)
res_AUC <- round(res_AUC, 2) # Digits
kable(res_AUC) %>%
  kable_styling("striped", position = "center") %>%
  add_header_above(c(" " = 1, 
                     "Apparent" = 3, 
                     "Internal" = 3, 
                     "Apparent + PGR" = 3, 
                     "Internal + PGR" = 3, 
                     "External" = 3, 
                     "External + PGR" = 3))

```

NOTE: add brief comments concordance
The time-dependent AUCs at 5 years were in the external validation were between 0.68 and 0.72 showing moderate discrimination. The additional information of the PGR marker increased the discrimination by 0.04.

The discrimination ability may be evaluated over the time since diagnosis through a plot showing the AUC over the time (years) in the development and validation set in the basic and extended model. Since the calculation of the AUC and the standard errors in the development set is computationally demanding, we showed only the plot in the validation data. However, we provided the R code (as a comment) for the development data.

```{r, plotAUC, fig.align='center'}

# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# NOTE: it takes long time to be computed
# Apparent without PGR
# AUC_rott5<-
#    timeROC(T=rott5$ryear, 
#            delta=rott5$rfs,
#            marker=predict(efit1,newdata=rott5),
#            cause=1,weighting='marginal',
#            times=quantile(rott5$ryear,probs=seq(0.01,0.85,0.02)),
#           iid=TRUE)

# Apparent with PGR
# AUC_rott5_pgr <-
#    timeROC(T=rott5$ryear, 
#            delta=rott5$rfs,
#            marker=predict(efit1_pgr,newdata=rott5),
#            cause=1,weighting='marginal',
#            times=quantile(rott5$ryear,probs=seq(0.01,0.85,0.02)),
#            iid=TRUE)

# Validation without PGR
AUC_gbsg5 <-
  timeROC(
    T = gbsg5$ryear, 
    delta = gbsg5$rfs,
    marker = predict(efit1, newdata = gbsg5),
    cause = 1, weighting = "marginal",
    times = quantile(gbsg5$ryear, probs = seq(0.02, 0.82, 0.02)),
    iid = TRUE
  )

# Validation with PGR
AUC_gbsg5_pgr <-
  timeROC(
    T = gbsg5$ryear, 
    delta = gbsg5$rfs,
    marker = predict(efit1_pgr, newdata = gbsg5),
    cause = 1, weighting = "marginal",
    times = quantile(gbsg5$ryear, probs = seq(0.02, 0.82, 0.02)),
    iid = TRUE
  )


# Calculate the confidence intervals
# AUC_rott5$lower <- AUC_rott5$AUC - qnorm(0.975) * AUC_rott5$inference$vect_sd_1
# AUC_rott5$upper <- AUC_rott5$AUC + qnorm(0.975) * AUC_rott5$inference$vect_sd_1
#
# AUC_rott5_pgr$lower <- AUC_rott5_pgr$AUC - qnorm(0.975)*AUC_rott5_pgr$inference$vect_sd_1
# AUC_rott5_pgr$upper <- AUC_rott5_pgr$AUC + qnorm(0.975) * AUC_rott5_pgr$inference$vect_sd_1

AUC_gbsg5$lower <- AUC_gbsg5$AUC - qnorm(0.975) * AUC_gbsg5$inference$vect_sd_1
AUC_gbsg5$upper <- AUC_gbsg5$AUC + qnorm(0.975) * AUC_gbsg5$inference$vect_sd_1

AUC_gbsg5_pgr$lower <- AUC_gbsg5_pgr$AUC - qnorm(0.975) * AUC_gbsg5_pgr$inference$vect_sd_1
AUC_gbsg5_pgr$upper <- AUC_gbsg5_pgr$AUC + qnorm(0.975) * AUC_gbsg5_pgr$inference$vect_sd_1


par(las = 1, xaxs = "i", yaxs = "i")
plot(AUC_gbsg5$times, 
     AUC_gbsg5$AUC,
     type = "l", 
     bty = "n",
     xlim = c(0, 5), 
     ylim = c(0, 1), 
     lwd = 2, 
     xlab = "Time (years)", 
     ylab = "AUC", 
     lty = 2
)
polygon(c(AUC_gbsg5$times, 
          rev(AUC_gbsg5$times)),
        c(AUC_gbsg5$lower, 
          rev(AUC_gbsg5$upper)),
  col = rgb(160, 160, 160, maxColorValue = 255, alpha = 100),
  border = FALSE
)
lines(AUC_gbsg5$times, 
      AUC_gbsg5$AUC, 
      col = "black", 
      lwd = 2, 
      lty = 2)
polygon(c(AUC_gbsg5_pgr$times, 
          rev(AUC_gbsg5_pgr$times)),
  c(AUC_gbsg5_pgr$lower, 
    rev(AUC_gbsg5_pgr$upper)),
  col = rgb(96, 96, 96, maxColorValue = 255, alpha = 100),
  border = FALSE
)
lines(AUC_gbsg5_pgr$times, 
      AUC_gbsg5_pgr$AUC, 
      col = "black", 
      lwd = 2, 
      lty = 1)
abline(h = 0.5)
legend("bottomright", 
       c("Original model", "Original model + PGR"), 
       lwd = 2, lty = c(2, 1), 
       bty = "n")
title("B GBSG data (n=686)", adj = 0)
```

The discrimination performance is higher in the extended model compared to the model without PGR over the entire follow-up.

### 2.3 Calibration
Calibration is the agreement between observed outcomes and predicted probabilities.
For example, in survival models, a predicted survival probability at a fixed time horizon _t_ of 80% is considered reliable if it can be expected that 80 out of 100 will survive among patients received a predicted survival probability of 80%.
Calibration can be assessed at a fixed time point (e.g. at 5 years), and globally (considering the entire range of the data).
In addition, different level of calibration assessment can be estimated according to the level of information available in the data. 

At a fixed time point t, mean calibration can be assessed using:
+ globally: considering the entire range of the data;

Since different level of information may be available, different level of calibration can be estimated:  mean, weak, and moderate calibration.

+ Mean calibration can be estimated:  

  + at a fixed time point:
    + using the Observed and Expected ratio at time t;  
    + using the difference between the log cumulative hazard from Cox model        with cloglog transformation of predicted probabilities as offset and         the mean of cloglog transformation.  
    
  + global:  
    + using the exponential of Poisson model intercept, exp(a)=(O/E), with         log of cumulative hazard as offset.  

+ Weak calibration can be estimated:
  + at a fixed time point  
    + using slope as the coefficient of cloglog transformation of             predicted probabilities in Cox model. Possible to determine intercept        after adjusting for slope.   
    
  + global:  
    + using slope as the coefficient of PI in Poisson model with log of            baseline cumulative hazard as offset.  

+ Moderate calibration can estimated:
  + at a fixed time point:  
    + using flexible calibration curve, complemented with ICI, E50, E90.  

  + global: 
    + using flexible calibration curve (requires development dataset or            baseline hazard at multiple time points).  

  

#### 2.3.1 Mean calibration
Please add some explanations. 

```{r, cal, message=FALSE, warning=FALSE}
# Apparent calibration measures (calibration in-the-large and calibration slope) without pgr
p <- log(predict(efit1, newdata = edata1, type = "expected") + .0000001)
lp <- predict(efit1, newdata = edata1, type = "lp")
logbase <- p - lp
rcfit1 <- glm(status ~ offset(p), family = poisson, data = edata1) # calibration in-the-large
rcfit2 <- glm(status ~ lp + offset(logbase), family = poisson, data = edata1)
# If we used the original data, the apparent internal validation may be also not perfect calibrated.


# External validation without pgr
p <- log(predict(efit1, newdata = evdata1, type = "expected") + .0000001)
lp <- predict(efit1, newdata = evdata1, type = "lp")
logbase <- p - lp
vcfit1 <- glm(status ~ offset(p), family = poisson, data = evdata1) # calibration in-the-large
vcfit2 <- glm(status ~ lp + offset(logbase), family = poisson, data = evdata1)

# Apparent calibration measures (calibration in-the-large and calibration slope) with pgr
p <- log(predict(efit1b, newdata = edata1, type = "expected") + .0000001)
lp <- predict(efit1b, newdata = edata1, type = "lp")
logbase <- p - lp
rcfit1b <- glm(status ~ offset(p), family = poisson, data = edata1) # calibration in-the-large
rcfit2b <- glm(status ~ lp + offset(logbase), family = poisson, data = edata1)


# External validation with pgr
p <- log(predict(efit1b, newdata = evdata1, type = "expected") + .0000001)
lp <- predict(efit1b, newdata = vdata, type = "lp")
logbase <- p - lp
vcfit1b <- glm(status ~ offset(p), family = poisson, data = evdata1) # calibration in-the-large
vcfit2b <- glm(status ~ lp + offset(logbase), family = poisson, data = evdata1)

# Results calibration in-the-large and slope
res_cal <- matrix(c(
  rcfit1$coefficients, rcfit1b$coefficients, vcfit1$coefficients, vcfit1b$coefficients,
  rcfit2$coefficients["lp"], rcfit2b$coefficients["lp"], vcfit2$coefficients["lp"], vcfit2b$coefficients["lp"]
),
nrow = 2, ncol = 4, byrow = T,
dimnames = list(
  c("Calibration in-the-large", "Calibration slope"),
  c("Apparent", "Apparent + PGR", "Validation", "Validation + PGR")
)
)
```

```{r, cal_res, echo=FALSE,message=FALSE,warning=FALSE}
res_cal <- matrix(c(
  rcfit1$coefficients,
  confint(rcfit1), # Apparent - calibration in-the-large

  rcfit1b$coefficients,
  confint(rcfit1b), # Apparent + PGR - calibration in-the-large

  vcfit1$coefficients,
  confint(vcfit1), # External - calibration in-the-large

  vcfit1b$coefficients,
  confint(vcfit1b), # External + PGR - calibration in-the-large

  rcfit2$coefficients["lp"],
  confint(rcfit2)[2, ], # Apparent - calibration slope

  rcfit2b$coefficients["lp"],
  confint(rcfit2b)[2, ], # Apparent + PGR - calibration slope

  vcfit2$coefficients["lp"],
  confint(vcfit2)[2, ], # External - calibration slope

  vcfit2b$coefficients["lp"],
  confint(vcfit2b)[2, ]
), # External + PGR - calibration slope
nrow = 2, ncol = 12, byrow = T,
dimnames = list(
  c("Calibration in-the-large", "Calibration slope"),
  rep(c("Estimate", "Lower .95 ", "Upper .95"), 4)
)
)

k <- 2
res_cal <- round(res_cal, k) # number of digits
kable(res_cal) %>%
  kable_styling("striped", position = "center") %>%
  add_header_above(c(" " = 1, "Apparent" = 3, "Apparent + PGR" = 3, "External" = 3, "External + PGR" = 3))
```

Calibration in-the-large was 0.13 and 0.10  and calibration slope was 1.01 and 1.12 in the validation data without and with PGR, respectively.
The exponential of calibration in-the-large is known as the observed and expected ratio (OE ratio). When the ratio is above 1, the model underpredicts the risk and below 1 indicates the overprediction of the model. The OE ratio is 1.14 and 1.11.

#### 2.3.2 Calibration plot using restricted cubic splines
Calibration plots of the external validation data with and without PGR are calculated and shown using restricted cubic splines.  
The interpretation of the calibration plot were provided in the section 2.3 of this document, in the corresponding paper and in the literature provided in the paper and at the end of this document. More details about the method are given in the references below.

```{r, cal_rcs, fig.align='center'}
## External data without PGR
evdata1$predmort5 <- 1 - survest(s5, newdata = evdata1, times = 5)$surv
# also the model (e.g. efit1) run using survival::coxph() may be used
# and 1-pec::predictSurvProb() function can replace 1-rms::survest()
evdata1$predmort5.cll <- log(-log(1 - evdata1$predmort5))

# Estimate
vcalibrate.efit1 <- cph(Surv(ryear, status) ~ rcs(predmort5.cll, 3),
  x = T, y = T,
  data = evdata1, surv = T
) # better rms::cph?
predict.grid <- seq(quantile(evdata1$predmort5, prob = 0.01), quantile(evdata1$predmort5, prob = 0.99), length = 100)
predict.grid.cll <- log(-log(1 - predict.grid))
predict.grid.df <- data.frame(predict.grid)
predict.grid.cll.df <- data.frame(predict.grid.cll)
names(predict.grid.df) <- "predmort5"
names(predict.grid.cll.df) <- "predmort5.cll"

# Plot
pred.vcalibrate.efit1 <- 1 - survest(vcalibrate.efit1, newdata = predict.grid.cll.df, times = 5)$surv
pred.vcalibrate.efit1.upper <- 1 - survest(vcalibrate.efit1, newdata = predict.grid.cll.df, times = 5)$lower
pred.vcalibrate.efit1.lower <- 1 - survest(vcalibrate.efit1, newdata = predict.grid.cll.df, times = 5)$upper
par(xaxs = "i", yaxs = "i", las = 1)
plot(predict.grid, pred.vcalibrate.efit1,
  type = "l", lty = 1, xlim = c(0, 1),
  ylim = c(0, 1), lwd = 2,
  xlab = "Predicted probability",
  ylab = "Observed probability", bty = "n"
)
lines(predict.grid, pred.vcalibrate.efit1.lower, type = "l", lty = 2, lwd = 2)
lines(predict.grid, pred.vcalibrate.efit1.upper, type = "l", lty = 2, lwd = 2)
abline(0, 1, lwd = 2, lty = 2, col = "red")
title("A External data without PGR", adj = 0)
par(new = T)
plot(density(evdata1$predmort5),
  axes = F, xlab = NA, ylab = NA,
  main = ""
)

## External data with PGR
evdata1$predmort5 <- 1 - survest(s6, newdata = evdata1, times = 5)$surv
evdata1$predmort5.cll <- log(-log(1 - evdata1$predmort5))

# Estimate
vcalibrate.efit1b <- cph(Surv(ryear, status) ~ rcs(predmort5.cll, 3),
  x = T, y = T, data = evdata1, surv = T
)
predict.grid <- seq(quantile(evdata1$predmort5, prob = 0.01), quantile(evdata1$predmort5, prob = 0.99), length = 100)
predict.grid.cll <- log(-log(1 - predict.grid))
predict.grid.df <- data.frame(predict.grid)
predict.grid.cll.df <- data.frame(predict.grid.cll)
names(predict.grid.df) <- "predmort5"
names(predict.grid.cll.df) <- "predmort5.cll"

# Plot
pred.vcalibrate.efit1b <- 1 - survest(vcalibrate.efit1b, newdata = predict.grid.cll.df, times = 5)$surv
pred.vcalibrate.efit1b.lower <- 1 - survest(vcalibrate.efit1b, newdata = predict.grid.cll.df, times = 5)$upper
pred.vcalibrate.efit1b.upper <- 1 - survest(vcalibrate.efit1b, newdata = predict.grid.cll.df, times = 5)$lower
par(xaxs = "i", yaxs = "i", las = 1)
plot(predict.grid, pred.vcalibrate.efit1b,
  type = "l", lty = 1, xlim = c(0, 1),
  ylim = c(0, 1), lwd = 2,
  xlab = "Predicted probability of recurrence at 5 years",
  ylab = "Observed probability of recurrence at 5 years", bty = "n"
)
lines(predict.grid, pred.vcalibrate.efit1b.lower, type = "l", lty = 2, lwd = 2)
lines(predict.grid, pred.vcalibrate.efit1b.upper, type = "l", lty = 2, lwd = 2)
abline(0, 1, lwd = 2, lty = 2, col = "red")
title("B External data with PGR", adj = 0)
par(new = T)
plot(density(evdata1$predmort5),
  axes = F, xlab = NA, ylab = NA,
  main = ""
)
```


Both plots identified good calibration although probabilities of recurrence were slightly underestimated especially for the lowest and the highest values of 
the observed probabilities of recurrence.  
The additional information of PGR improved the overall calibration, especially for the highest values, as shown in the two calibration plots above.


#### 2.3.3 Other calibration metrics: ICI, E50,E90 and Emax based on restricted cubic splines
As explained in the previous sections, three additional calibration metrics were provided: the integrated Calibration Index (ICI), E50, E90 and Emax.  
ICI is the weighted difference between smoothed observed proportions and predicted probabilities in which observations are weighted by the empirical density function of the predicted probabilities.   
E50, E90 and Emax denote the median, the 90th percentile and the maximum absolute difference between observed and predicted probabilities of the outcome at time *t*. 

```{r, cal_rcs_metrics, fig.align='center'}
# Calibration metrics

# Apparent
edata1$predmort5 <- 1 - predictSurvProb(efit1, newdata = edata1, times = 5)
# Here we used the model created with  survival::coxph() (e.g.efit1) but again 1-survest() using the model created with rms:cph() (e.g s5) can be used
edata1$predmort5.cll <- log(-log(1 - edata1$predmort5))
rcalibrate.efit1 <- coxph(Surv(ryear, status) ~ rcs(predmort5.cll, 3), x = T, y = T, data = edata1) # rms:cph() can be also used
pred.rcalibrate.efit1 <- 1 - predictSurvProb(rcalibrate.efit1,
  newdata = edata1, times = 5
)

ICI.edata1.5yr <- mean(abs(edata1$predmort5 - pred.rcalibrate.efit1))
E50.edata1.5yr <- median(abs(edata1$predmort5 - pred.rcalibrate.efit1))
E90.edata1.5yr <- quantile(abs(edata1$predmort5 - pred.rcalibrate.efit1), probs = 0.9)
Emax.edata1.5yr <- max(abs(edata1$predmort5 - pred.rcalibrate.efit1))

# Apparent + PGR
edata1$predmort5 <- 1 - predictSurvProb(efit1b, newdata = edata1, times = 5)
edata1$predmort5.cll <- log(-log(1 - edata1$predmort5))
rcalibrate.efit1b <- coxph(Surv(ryear, status) ~ rcs(predmort5.cll, 3), x = T, y = T, data = edata1) # better rms::cph?
pred.rcalibrate.efit1b <- 1 - predictSurvProb(rcalibrate.efit1b,
  newdata = edata1, times = 5
)

ICI.edata1b.5yr <- mean(abs(edata1$predmort5 - pred.rcalibrate.efit1b))
E50.edata1b.5yr <- median(abs(edata1$predmort5 - pred.rcalibrate.efit1b))
E90.edata1b.5yr <- quantile(abs(edata1$predmort5 - pred.rcalibrate.efit1b), probs = 0.9)
Emax.edata1b.5yr <- max(abs(edata1$predmort5 - pred.rcalibrate.efit1b))

# Validation
evdata1$predmort5 <- 1 - predictSurvProb(efit1, newdata = evdata1, times = 5)
evdata1$predmort5.cll <- log(-log(1 - evdata1$predmort5))
vcalibrate.efit1 <- coxph(Surv(ryear, status) ~ rcs(predmort5.cll, 3), x = T, y = T, data = evdata1) # better rms::cph?
pred.vcalibrate.efit1 <- 1 - predictSurvProb(vcalibrate.efit1,
  newdata = evdata1, times = 5
)

ICI.evdata1.5yr <- mean(abs(evdata1$predmort5 - pred.vcalibrate.efit1))
E50.evdata1.5yr <- median(abs(evdata1$predmort5 - pred.vcalibrate.efit1))
E90.evdata1.5yr <- quantile(abs(evdata1$predmort5 - pred.vcalibrate.efit1), probs = 0.9)
Emax.evdata1.5yr <- max(abs(evdata1$predmort5 - pred.vcalibrate.efit1))

# Validation + PGR
evdata1$predmort5 <- 1 - predictSurvProb(efit1b, newdata = evdata1, times = 5)
evdata1$predmort5.cll <- log(-log(1 - evdata1$predmort5))
vcalibrate.efit1b <- coxph(Surv(ryear, status) ~ rcs(predmort5.cll, 3), x = T, y = T, data = evdata1) # better rms::cph?
pred.vcalibrate.efit1b <- 1 - predictSurvProb(vcalibrate.efit1b,
  newdata = evdata1, times = 5
)

ICI.evdata1b.5yr <- mean(abs(evdata1$predmort5 - pred.vcalibrate.efit1b))
E50.evdata1b.5yr <- median(abs(evdata1$predmort5 - pred.vcalibrate.efit1b))
E90.evdata1b.5yr <- quantile(abs(evdata1$predmort5 - pred.vcalibrate.efit1b), probs = 0.9)
Emax.evdata1b.5yr <- max(abs(evdata1$predmort5 - pred.vcalibrate.efit1b))
```

```{r, calmetrics_res, echo=FALSE}

res_calmetrix <- matrix(c(
  ICI.edata1.5yr, ICI.edata1b.5yr, ICI.evdata1.5yr, ICI.evdata1b.5yr,
  E50.edata1.5yr, E50.edata1b.5yr, E50.evdata1.5yr, E50.evdata1b.5yr,
  E90.edata1.5yr, E90.edata1b.5yr, E90.evdata1.5yr, E90.evdata1b.5yr,
  Emax.edata1.5yr, Emax.edata1b.5yr, Emax.evdata1.5yr, Emax.evdata1b.5yr
),
byrow = T, ncol = 4, nrow = 4
)
colnames(res_calmetrix) <- c("Apparent", "Apparent + PGR", "External", "External + PGR")
rownames(res_calmetrix) <- c("ICI", "E50", "E90", "Emax")

k <- 2
res_calmetrix <- round(res_calmetrix, k) # number of digits
kable(res_calmetrix) %>% kable_styling("striped", position = "center")
```

In the validation data, PGR improves slightly the ICI and the median of the absolute difference between observed and predicted probabilities at 5 years.  More imprecise absolute differences between and observed and predicted probabilities in the extreme values of the distribution (E90, Emax). This is also graphically confirmed by the calibration plots.


## Goal 3 - Clinical utility
Discrimination and calibration measures are essential to assess the prediction performance but insufficient to evaluate the potential clinical utility of a risk prediction model for decision making. When new markers are available, clinical utility assessment evaluates whether the extended model helps to improve decision making.  
Clinical utility is measured by the net benefit that includes the number of true positives and the number of false positives. For example, in time-to-event models, the true positives reflect the benefit of being event free for a given time horizon using additional interventions such as additional treatments, personalized follow-up or additional surgeries. The false positives represent the harms of unnecessary interventions.   
Generally, in medicine, clinicians accepts to treat a certain number of patients for which interventions are unnecessary to be event free for a given time horizon. So, false negatives (the harm of not being event free for a given time horizon) are more important than false positives (the harm of unnecessary interventions). Thus, net benefit is the number of true positives classifications minus the false positives classifications weighted by a factor related to the harm of not preventing the event versus unnecessary interventions. The weighting is derived from the threshold probability to death (one minus survival probability) using a defined time horizon (for example 5 years since diagnosis). For example, a threshold of 10% implies that additional interventions for 10 patients of whom one would have experience the event in 5 years if untreated is acceptable (thus treating 9 unnecessary patients). This strategy is compared with the strategies of treat all and treat none patients. If overtreatment is harmful, a higher threshold should be used.  
The net benefit is calculated as:  
  
<img src="https://render.githubusercontent.com/render/math?math=%5Chuge%7B%5Cfrac%7BTP%7D%7Bn%7D-%5Cfrac%7BFP%7D%7Bn%7D*%5Cfrac%7Bp_t%7D%7B1-p_t%7D%7D">
  
*TP*=true positive patients   
*FP*=false positive patients  
*n*=number of patients and *p*<sub>t</sub> is the risk threshold.  

For survival data *TP* and *FP* is calculated as follows:   
<img src="https://render.githubusercontent.com/render/math?math=%5CLarge%7BTP%20%3D%20%5B1-S(t)%7C%20X%3D1%5D*P(X%3D1)*n%7D">
  
<img src="https://render.githubusercontent.com/render/math?math=%5CLarge%7BFP%20%3D%20%5BS(t)%7C%20X%3D1%5D*P(X%3D1)*n%7D">
  
  where  
*S(t)* survival at time *t*  
  *X=1* where the predicted probability at time *t* is  *p*<sub>t</sub>  
  
  And the the decision curve is calculated as follows:
  
1. Choose a time horizon (in this case 5 years);
2. Specify a risk threshold which reflects the ratio between harms and benefit of an additional intervention;
3. Calculate the number of true positive and false positive given the threshold specified in (2);
4. Calculate the net benefit of the survival model;
5. Plot net benefit on the *y-axis* against the risk threshold on the *x-axis*;
6. Repeat steps 2-4 for each model consideration;
7. Repeat steps 2-4 for the strategy of assuming all patients are treated;
8. Draw a straight line parallel to the *x-axis* at y=0 representing the net benefit associated with the strategy of assuming that all patients are not treated.

Given some thresholds, the model/strategy with higher net benefit represents the one that potentially improves  clinical decision making. However, poor discrimination and calibration lead to lower net benefit.

```{r, function_stdca, message=FALSE,warning=FALSE, fig.align='center',include=FALSE}
# Run the function to calculate the net benefit and the elements needed to develop decision curve analysis
source(here::here('Functions/stdca.R'))
```


```{r, dca, message=FALSE,warning=FALSE, fig.align='center'}
# We could use the rms::survest() or pec::predictEventProb() to get the mortality at observed time
# Development data
# Predicted probability calculation
edata1$mort5_efit1 <- 1 - predictSurvProb(efit1, newdata = edata1, times = 5)

# Extended model with PGR
# Predicted probability calculation
edata1$mort5_efit1b <- 1 - predictSurvProb(efit1b, newdata = edata1, times = 5)

# Run decision curve analysis

# Development data
# Model without PGR
edata1 <- as.data.frame(edata1)
dca_rdata_1 <- stdca(
  data = edata1, outcome = "status", ttoutcome = "ryear",
  timepoint = 5, predictors = "mort5_efit1", xstop = 1.0,
  ymin = -0.01, graph = FALSE
)
# Model with PGR
dca_rdata_1b <- stdca(
  data = edata1, outcome = "status", ttoutcome = "ryear",
  timepoint = 5, predictors = "mort5_efit1b", xstop = 1.0,
  ymin = -0.01, graph = FALSE
)

# Decision curves plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(dca_rdata_1$net.benefit$threshold,
  dca_rdata_1$net.benefit$mort5_efit1,
  type = "l", lwd = 2, lty = 1,
  xlab = "Threshold probability in %", ylab = "Net Benefit",
  xlim = c(0, 1), ylim = c(-0.10, 0.45), bty = "n",
  cex.lab = 1.2, cex.axis = 1
)
# legend('topright',c('Treat all','Treat none','Prediction model'),
#        lwd=c(2,2,2),lty=c(1,1,2),col=c('darkgray','black','black'),bty='n')
lines(dca_rdata_1$net.benefit$threshold, dca_rdata_1$net.benefit$none, type = "l", lwd = 2, lty = 4)
lines(dca_rdata_1$net.benefit$threshold, dca_rdata_1$net.benefit$all, type = "l", lwd = 2, col = "darkgray")
lines(dca_rdata_1b$net.benefit$threshold, dca_rdata_1b$net.benefit$mort5_efit1b, type = "l", lwd = 2, lty = 5)
legend("topright",
  c(
    "Treat All",
    "Original model",
    "Original model + PGR",
    "Treat None"
  ),
  lty = c(1, 1, 5, 4), lwd = 2, col = c("darkgray", "black", "black", "black"),
  bty = "n"
)
title("A Development data", adj = 0, cex = 1.5)


# External data
# Validation data
# Predicted probability calculation
evdata1$mort5_model1 <- 1 - predictSurvProb(efit1, newdata = evdata1, times = 5)

# Extended model with PGR
# Predicted probability calculation
evdata1$mort5_model1b <- 1 - predictSurvProb(efit1b, newdata = evdata1, times = 5)

# Run decision curve analysis

# Development data
# Model without PGR
evdata1 <- as.data.frame(evdata1)
dca_vdata_model1 <- stdca(
  data = evdata1, outcome = "status", ttoutcome = "ryear",
  timepoint = 5, predictors = "mort5_model1", xstop = 1.0,
  ymin = -0.01, graph = FALSE
)
# Model with PGR
dca_vdata_model1b <- stdca(
  data = evdata1, outcome = "status", ttoutcome = "ryear",
  timepoint = 5, predictors = "mort5_model1b", xstop = 1,
  ymin = -0.01, graph = FALSE
)

# Decision curves plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(dca_vdata_model1$net.benefit$threshold,
  dca_vdata_model1$net.benefit$mort5_model1,
  type = "l", lwd = 2, lty = 1,
  xlab = "Threshold probability in %", ylab = "Net Benefit",
  xlim = c(0, 1), ylim = c(-0.10, 0.60), bty = "n",
  cex.lab = 1.2, cex.axis = 1
)
# legend('topright',c('Treat all','Treat none','Prediction model'),
#        lwd=c(2,2,2),lty=c(1,1,2),col=c('darkgray','black','black'),bty='n')
lines(dca_vdata_model1$net.benefit$threshold, dca_vdata_model1$net.benefit$none, type = "l", lwd = 2, lty = 4)
lines(dca_vdata_model1$net.benefit$threshold, dca_vdata_model1$net.benefit$all, type = "l", lwd = 2, col = "darkgray")
lines(dca_vdata_model1b$net.benefit$threshold, dca_vdata_model1b$net.benefit$mort5_model1b, type = "l", lwd = 2, lty = 5)
legend("topright",
  c(
    "Treat All",
    "Original model",
    "Original model + PGR",
    "Treat None"
  ),
  lty = c(1, 1, 5, 4), lwd = 2, col = c("darkgray", "black", "black", "black"),
  bty = "n"
)
title("B External data", adj = 0, cex = 1.5)
```

Based on previous research we used a range of thresholds from 14% to 23% for adjuvant chemotherapy. If we choose a threshold of 20% the model had a net benefit of 0.262 in the development data using the basic model. This means that the model would identify 26 patients per 100 who will have recurrent breast cancer or die within 5 years since diagnosis and thus adjuvant chemotherapy is really needed.   
The decision curve shows that the net benefit would be much larger for higher threshold values, i.e., patients accepting higher risks of recurrence compared to the treat all strategy. The same interpretation may be used in the validation data:
  choosing a threshold of 20% the basic model had a net benefit of 0.385 for the basic and the extended model.  
Moreover, net benefit can be defined in terms of reduction of avoidable interventions (e.g adjuvant chemotherapy per 100 patients) by:
  
<img src="https://render.githubusercontent.com/render/math?math=%5Chuge%7B%5Cfrac%7BNB_%7Bmodel%7D%20-%20NB_%7Ball%7D%7D%7B(p_t%2F%20(1-p_t))%7D*100%7D%0A">  
  
  where *NB*<sub>model</sub> is the net benefit of the prediction model, *NB*<sub>all</sub> is the net benefit of the strategy treat all and $p_{t}$ is the risk threshold.



## References
+ Overall measures   
  Reference: https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-018-0029-2*/ \
  R Vignette: https://cran.r-project.org/web/packages/riskRegression/vignettes/IPA.html#fn.1 \

+ Discrimination measures \ 
  https://www.jstor.org/stable/27639883 \
  https://onlinelibrary.wiley.com/doi/10.1002/sim.5958 \
  https://pubmed.ncbi.nlm.nih.gov/23172755/ \
  

+ Calibration \
  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3933449/pdf/nihms542648.pdf \
  https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8281   \
  https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8570   \

+ Clinical utility (decision curves)  
  R/SAS/STATA code and references: 
  https://www.mskcc.org/departments/epidemiology-biostatistics/biostatistics/decision-curve-analysis \
  More guidelines about net benefit assessment and interpretation \
  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6261531/ \
  https://diagnprognres.biomedcentral.com/articles/10.1186/s41512-019-0064-7 \
  
+ Other useful references \
  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6728752/ \
  https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7100774/ \
  
## Reproducibility ticket

```{r repro_ticket, echo=TRUE}
sessioninfo::session_info()
```
  
  
  
