---
title: "Performance assessment of survival prediction models - extended code"
always_allow_html: true
output: 
  github_document:
    toc: true
    toc_depth: 5
  keep_text: true
  pandoc_args: --webtex
---
```{r setup, include=FALSE}
# Knitr options
knitr::opts_chunk$set(
  fig.retina = 3,
  fig.path = "imgs/03_predsurv_extended/",
  echo = FALSE
)
```

## Goals
In this document, we assume that individual data of the development and validation set are both available.

In particular:   
1. To develop a risk prediction model with a time-to-event outcome;  
2. To assess the prediction performance of a model with a time-to-event outcome;  
3. To assess the potential clinical utility of a risk prediction model with time-to-event outcome.

The extended code basically evaluates the prediction performance at a fixed time horizon (e.g. 5 years) and also for the entire follow-up time.

### Load packages and import data
We following libraries are needed to achieve the following goals, if you have not them installed, please use
install.packages('') (e.g. install.packages('survival')) or use the user-friendly approach if you are using RStudio.

```{r, wdlib, message=FALSE,warning=FALSE}

# Use pacman to check whether packages are installed, if not load
if (!require("pacman")) install.packages("pacman")
library(pacman)

pacman::p_load(
  rio,
  survival,
  rms,
  pec,
  riskRegression,
  timeROC,
  plotrix,
  splines,
  knitr,
  kableExtra,
  gtsummary,
  boot,
  tidyverse,
  rsample,
  gridExtra,
  webshot
)
```

The two primary datasets from the Rotterdam study and the
German Breast Cancer Study Group (GBSG) are 
included as example data sets in the ```survival``` package.
The Rotterdam data has separate variables for time to recurrence
of disease and time to death, and these are combined into a single
endpoint of recurrence free survival (RFS), while the gbsg data
set has RFS as its single endpoint.
The follow-up time is converted from days to years, that being a more
natural unit for any plots; it does not change any fits.
A categorical variable for nodes is also added.


### Data preparation and descriptive statistics

<details>
  <summary>Click to expand code</summary>
```{r, import,fig.align='center', echo=TRUE, eval=FALSE}
# Development data
# Recurrence free survival is the time until the earlier of
#  recurrence or death. 
rotterdam$ryear <- rotterdam$rtime/365.25  # time in years
rotterdam$rfs <- with(rotterdam, pmax(recur, death)) #The variable rfs is a status indicator, 0=alive without relapse, 1= death or relapse.
rotterdam$ryear[rotterdam$rfs == 1 & rotterdam$recur == 0 & rotterdam$death == 1 & (rotterdam$rtime < rotterdam$dtime)] <- rotterdam$dtime[rotterdam$rfs == 1 & rotterdam$recur == 0 & rotterdam$death == 1 & (rotterdam$rtime < rotterdam$dtime)]/365.25  #Fix the outcome for 43 patients who have died but censored at time of recurrence which was less than death time. The actual death time should be used rather than the earlier censored recurrence time.

# variables used in the analysis
pgr99 <- quantile(rotterdam$pgr, .99, type = 1) # there is a large outlier of 5000
rotterdam$pgr2 <- pmin(rotterdam$pgr, pgr99) # Winsorized value
rotterdam$csize <- rotterdam$size           # categorized size
rotterdam$cnode <- cut(rotterdam$nodes, c(-1, 0, 3, 51),
                       c("0", "1-3", ">3"))   # categorized node
rotterdam$grade3 <- as.factor(rotterdam$grade)
levels(rotterdam$grade3) <- c("1-2", "3")

# Save in the data the restricted cubic spline term using Hmisc::rcspline.eval() package
rcs3_pgr <- rcspline.eval(rotterdam$pgr2, knots = c(0, 41, 486))
attr(rcs3_pgr, "dim") <- NULL
attr(rcs3_pgr, "knots") <- NULL
rotterdam$pgr3 <- rcs3_pgr

# Validation data
gbsg$ryear <- gbsg$rfstime/365.25
gbsg$rfs   <- gbsg$status           # the GBSG data contains RFS
gbsg$cnode <- cut(gbsg$nodes, c(-1,0, 3, 51),
                       c("0", "1-3", ">3"))   # categorized node
gbsg$csize <- cut(gbsg$size,  c(-1, 20, 50, 500), #categorized size
                  c("<=20", "20-50", ">50"))
gbsg$pgr2 <- pmin(gbsg$pgr, pgr99) # Winsorized value
gbsg$grade3 <- as.factor(gbsg$grade)
levels(gbsg$grade3) <- c("1-2", "1-2", "3")

# Restricted cubic spline for PGR
rcs3_pgr <- rcspline.eval(gbsg$pgr2, knots = c(0, 41, 486))
attr(rcs3_pgr, "dim") <- NULL
attr(rcs3_pgr, "knots") <- NULL
gbsg$pgr3 <- rcs3_pgr


# Much of the analysis will focus on the first 5 years: create
#  data sets that are censored at 5
temp <- survSplit(Surv(ryear, rfs) ~ ., data = rotterdam, 
                  cut = 5,
                  episode = "epoch")
rott5 <- subset(temp, epoch == 1)  # only the first 5 years
temp <- survSplit(Surv(ryear, rfs) ~ ., data = gbsg, 
                  cut = 5,
                  episode = "epoch")
gbsg5 <- subset(temp, epoch == 1)

# Relevel
rott5$cnode <- relevel(rotterdam$cnode, "0")
gbsg5$cnode <- relevel(gbsg$cnode, "0")
```
 </details>
  
```{r, import, fig.align='center', warning=FALSE, eval=TRUE}
```


```{r tab1_bis, echo=FALSE}
rsel <- rotterdam[, c("pid", "csize", "cnode", "grade3", "age", "pgr")]
vsel <- gbsg[, c("pid", "csize", "cnode", "grade3", "age", "pgr")]
rsel$dt <- 1
vsel$dt <- 2
cdata <- rbind(rsel, vsel)
cdata$dt <- factor(cdata$dt,
  levels = c(1, 2),
  labels = c("Development dataset", "Validation dataset")
)

label(cdata$csize) <- "Size"
label(cdata$cnode) <- "Number of nodes"
label(cdata$grade3) <- "Grade of tumor"
label(cdata$age) <- "Age"
label(cdata$pgr) <- "PGR"
label(cdata$dt) <- "Dataset"

units(cdata$csize) <- "mm"
units(cdata$age) <- "years"
units(cdata$pgr) <- "ng/mL"


gtsummary::tbl_summary(
  data = cdata |> select(-pid),
  label = list(age ~ "Age (years)", csize ~ "Size (cm)", 
               pgr ~ "PGR (ng/mL)",
               grade3 ~ "Grade"),
  by = "dt", 
  type = all_continuous() ~ "continuous2",
  statistic = all_continuous() ~ c(
    "{mean} ({sd})",
    "{median} ({min}, {max})"
  ),
) |> 
  gtsummary::as_kable_extra() |> 
  kableExtra::kable_styling("striped") 
```


## Goal 1 - Develop a risk prediction model with a time-to-event outcome
Prediction models are useful to provide the estimated probability of a specific outcome using personal information.
In many studies, especially in medicine, the main outcome under assessment is the time to an event of interest defined generally as survival time. Prognostic models for survival end points, such as recurrence or progression of disease, need to account for drop out during follow-up. Patients who have not experienced the event of interest are censored observations. Cox regression analysis is the most popular statistical model to deal with such data in oncology and other medical research.  

### 1.1 Preliminary investigation - survival and censoring curves in the development and validation data
First, we draw the survival and the censoring curves of the development and validation data

<details>
  <summary>Click to expand code</summary>
```{r, surv, fig.align='center', echo=TRUE, eval=FALSE}
# Development set
sfit_rott <- survfit(Surv(ryear, rfs == 1) ~ 1, 
                     data = rotterdam) # survival
sfit_rott_c <- survfit(Surv(ryear, rfs == 0) ~ 1, 
                       data = rotterdam) # censoring

par(xaxs = "i", yaxs = "i", las = 1)
plot(sfit_rott, 
     conf.int = FALSE, 
     lwd = 2, 
     xlab = "Years", 
     bty = "n")
lines(sfit_rott_c, 
      conf.int = FALSE, 
      col = 2, 
      lwd = 2)
legend(11, .9, 
       c("Death", "Censoring"), 
       col = 1:2, 
       lwd = 2, 
       bty = "n")
title("Development set")


# Validation set
sfit_gbsg <- survfit(Surv(ryear, rfs == 1) ~ 1, 
                     data = gbsg) # survival
sfit_gbsg_c <- survfit(Surv(ryear, rfs == 0) ~ 1, 
                       data = gbsg) # censoring

par(xaxs = "i", yaxs = "i", las = 1)
plot(sfit_gbsg, 
     conf.int = FALSE, 
     lwd = 2, 
     xlab = "Years", 
     bty = "n", 
     xlim = c(0, 8))
lines(sfit_gbsg_c, 
      conf.int = FALSE, 
      col = 2,
      lwd = 2)
legend("bottomleft", 
       c("Death", "Censoring"), 
       col = 1:2, 
       lwd = 2, 
       bty = "n")
title("Validation set")
```
 </details>
 
```{r, surv, fig.align='center', warning=FALSE, eval=TRUE}
```

A number of 2982 patients were included to develop the risk prediction model for survival with a median follow-up of 9 years. The 5-year survival was 56% (95% CI: 55-58%). 
A number of 686 patients were selected to externally validate the risk prediction model. The median survival in the validation data was 4.5 years. The median survival was 5 years while the 5-year survival was 49% (95% CI: 45-54%).


### 1.2 Secondary investigation - check non-linearity of continuous predictors
The potential non-linear relation between continuous predictors (i.e. progesterone level) and the outcome should be investigated before developing a risk prediction model. Non-linearity of continuous predictors can be checked using splines.  
Physically, a spline is a flexible wood or metal strip, which is passed through a set of fixed points (knots) in order to approximate a curve. The most common computational approximation to this is a cubic smoothing spline which is cubic between the knot points, and constrained to be linear beyond the two end knots. 
For the restricted cubic spline using `rms::rcs()` R package::function(), the position of the knots are defined at 10<sup>th</sup>,50<sup>th</sup> and 90<sup>th</sup> quantile of the continuous predictor distribution. For more details see Frank Harrell's book 'Regression Model Strategies' on page 27 (second edition).  
The user can specify the positions of the knots instead of using the default calculation of the knots proposed in the book of Frank Harrell.
To deal with very large influential value, we winzorize progesterone level to the 90<sup>th</sup>  percentile.


<details>
  <summary>Click to expand code</summary>
```{r,ff, warning=FALSE, fig.align='center', echo=TRUE, eval=FALSE}
dd <- datadist(rotterdam)
options(datadist = "dd")
fit_pgr <- cph(Surv(ryear, rfs) ~ rcs(pgr2), 
               data = rotterdam, x = T, y = T, surv = T)
plot(Predict(fit_pgr))
options(datadist = NULL)
``` 
</details>
  
```{r, ff, fig.align='center', warning=FALSE, eval=TRUE}
```

We should model the progesterone level using a three-knot restricted cubic spline. We save the spline in the development and validation data.


### 1.3 Model development - first check - the proportional hazard (PH) assumption
We now examine the fits in a more careful way by checking the proportionality of the hazards of the Cox regression model.
Firstly, we fit the first prediction model in the development data using size, node, grade. Then, we check the PH assumption.

<details>
  <summary>Click to expand code</summary>
```{r,ph, message=FALSE, warning=FALSE,fig.align='center', echo=TRUE, eval=FALSE}
fit1_ph <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3, 
                data = rotterdam, x = T, y = T)


zp1 <- cox.zph(fit1_ph, transform = "identity")
kable(round(zp1$table, 3)) |> kable_styling("striped", position = "center")

oldpar <- par(mfrow = c(2, 2), mar = c(5, 5, 1, 1))
for (i in 1:3) {
  plot(zp1[i], resid = F)
  abline(0, 0, lty = 3)
}
par(oldpar)
```
 </details>
  
```{r, ph, fig.align='center', warning=FALSE, eval=TRUE}
```

The statistical tests show strong evidence of non-proportionality. Since the number of death is large the formal tests are quite sensitive, however, and it is important to also examine the graphs.  
These show an estimated coefficient as a function of time. As a further follow-up we will divide the data into 3 epochs of 0-5, 5-10, and 10+ years, fitting a separate model to each.

<details>
  <summary>Click to expand code</summary>
```{r, epoch, warning=FALSE, echo=TRUE, eval=FALSE}
# Development
edata <- survSplit(Surv(ryear, rfs) ~ .,
  data = rotterdam, cut = c(5, 10),
  episode = "epoch"
)
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = edata[edata$epoch == 1, ], x = T, y = T
)
efit2 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = edata[edata$epoch == 2, ], x = T, y = T
)
efit3 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = edata[edata$epoch == 3, ], x = T, y = T
)
```
</details>
  
```{r, epoch, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r,ph_table, echo=FALSE}
res_efit <- round(rbind(e1 = coef(efit1), 
                        e2 = coef(efit2), 
                        e3 = coef(efit3)), 2)
rownames(res_efit) <- c("Epoch 1: 0-5 years", 
                        "Epoch 2: 5-10 years", 
                        "Epoch 3: >10 years")
kable(res_efit) |> kable_styling("striped", position = "center")

options(prType = "html")
tt1 <- table(edata$epoch, edata$rfs, dnn = c("Epoch", "Status"))
rownames(tt1) <- c("Epoch 1: 0-5 years", 
                   "Epoch 2: 5-10 years", 
                   "Epoch 3: >10 years")
kable(tt1,
  col.names = c("Censored", "Event"),
  row.names = TRUE
) |> kable_styling("striped", position = "center")
```

A drastic change in the size coefficients across all epochs is apparent, along with a major reduction in the nodes coefficient in epoch 3. As an ameleoration of this we will refit the model using only the first epoch, which includes most of the recurrences and deaths.  
We applied the administrative censoring at 5 years in the development data and we assessed the prediction performance of the prognostic model at 5 years. The 5-year horizon was chosen because it is also a common prediction time horizon in clinical practice.  The hazards in the development data seem not totally proportional within 5 years but minor deviation of proportionality were considered acceptable. 

### 1.4 Model development - fit the risk prediction models
We develop the risk prediction model in the development data considering the first 5-year follow-up to minimize the violation of proportional hazard including size, nodel and grade. The second model also includes the progesterone level modelled using a 3-knot restricted cubic spline.  
We also administratively censored the validation data at 5 years.

<details>
  <summary>Click to expand code</summary>
```{r, model_development, fig.align='center', echo=TRUE, eval=FALSE}
# Consider the first 5-year epoch in the development set
# Refit the model
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Baseline at 5 years
bh <- basehaz(efit1, centered = FALSE) # uncentered
bh$surv <- exp(-bh$hazard) # baseline survival
S0_t5 <- bh$surv[bh$time == 5] 
# NOTE: this can be used to calculate S(t = 5) = S0(t = 5)**exp(X*beta)

bh_c <- basehaz(efit1, centered = TRUE)
bh_c$surv_c <- exp(-bh_c$hazard) # baseline survival
S0_t5_c <- bh_c$surv[bh_c$time == 5] 
```
</details>
  
```{r, model_development, fig.align='center', warning=FALSE, eval=TRUE}
```

Below the results of the models:  

+ Classical model:  
```{r, summary_m1, fig.align='center',warning=FALSE, echo = FALSE}
dd <- datadist(rott5)
options(datadist = "dd")
options(prType = "html")
fit_cph_rott5 <- cph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T, surv = T
)
print(fit_cph_rott5)
```

+ Extended model:  

```{r, summary_m2, fig.align='center',warning=FALSE , echo = FALSE}
options(prType = "html")
fit_cph_rott5_pgr <- update(fit_cph_rott5, . ~ . 
                            + rcs(pgr2, c(0, 41, 486)))
print(fit_cph_rott5_pgr)
options(datadist = NULL)
```

The coefficients of the models indicated that higher size, higher number of positive lymph nodes and higher grade is more associate with poorer prognosis. The association of the progesterone biomarker and the outcome is non-linear as investigated previously.


## Goal 2 - Assessing performance in survival prediction models
The performance of a risk prediction models may be evaluated through:  

+ discrimination: the ability of the model to correctly rank patients with and without the outcome by a certain time point. This requires the coefficients (or the log of the hazard ratios) of the developed Cox prediction model to be evaluated;     

+ calibration: the agreement between observed and predicted probabilities. It additionally requires the baseline (cumulative) hazard or survival;     
  
+ overall performance measures: a combination of discrimination and calibration.  
  
Unfortunately, few publications report the complete baseline (cumulative) hazard or survival or even the baseline (cumulative) hazard or survival at fixed time horizon *t*.  
It is common that physicians focus on one or more clinically relevant time horizons to inform subjects about their risk. 
We aim to assess the prediction performance of a risk prediction model with time-to-event outcome in a new data when information at a fixed time horizon(s) (here at 5 years) of a developed prediction model were provided. 

### 2.1 Discrimination measures
Discrimination is the ability to differentiate between subjects who have the outcome by a certain time point and subjects who do not.
Concordance can be assessed over several different time intervals:

+ the entire range of the data. Two concordance measures are suggested:    

  + Harrell's C quantifies the degree of concordance as the proportion of evaluable pairs where the patient with a longer survival time has better predicted survival;  
  
  + Uno's C uses a time dependent weighting that more fully adjusts for censoring;  
  
+ a 5 year window corresponding to our target assessment point. Uno's cumulative/dynamic time-dependent Area Under the Curve (AUC) is suggested. Uno's time-dependent AUC summarizes discrimination at specific fixed time points. At any time point of interest, _t_, a patient is classified as having an event if the patient experienced the event between baseline and _t_ (5 years in our case study), and as a non-event if the patient remained event-free at _t_. The time-dependent AUC evaluates whether predicted probabilities were higher for cases than for non-cases.

There is some uncertainty in the literature about the original Harrell
formulation versus Uno's suggestion to re-weight the time scale by the
factor $1/G^2(t)$ where $G$ is the censoring distribution.
There is more detailed information in the concordance vignette found in the
survival package.

For all three measures, values close to 1 indicate good discrimination ability, while values close to 0.5 indicated poor discrimination ability.  

<details>
  <summary>Click to expand code</summary>
```{r, concordance, message=FALSE ,warning=FALSE, echo=TRUE, eval=FALSE}

# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

## Save elements needed to estimate concordance indexes
# Development data 
rott5$lp <- predict(efit1, newdata = rott5)
rott5$lp_pgr <- predict(efit1_pgr, newdata = rott5)

# Validation data 
gbsg5$lp <- predict(efit1, newdata = gbsg5)
gbsg5$lp_pgr <- predict(efit1_pgr, newdata = gbsg5)


# Harrell C - development
harrell_C_rott5 <- concordance(Surv(ryear, rfs) ~ lp, 
                               rott5, 
                               reverse = TRUE)

harrell_C_rott5_pgr <- concordance(Surv(ryear, rfs) ~ lp_pgr, 
                               rott5, 
                               reverse = TRUE)

# Harrell C - validation
harrell_C_gbsg5 <- concordance(Surv(ryear, rfs) ~ lp, 
                               gbsg5, 
                               reverse = TRUE)

harrell_C_gbsg5_pgr <- concordance(Surv(ryear, rfs) ~ lp_pgr, 
                               gbsg5, 
                               reverse = TRUE)

# Uno's C - development
Uno_C_rott5 <- concordance(Surv(ryear, rfs) ~ lp, 
                           rott5, 
                           reverse = TRUE,
                           timewt = "n/G2")

Uno_C_rott5_pgr <- concordance(Surv(ryear, rfs) ~ lp_pgr, 
                           rott5, 
                           reverse = TRUE,
                           timewt = "n/G2")

# Uno's C - validation
Uno_C_gbsg5 <- concordance(Surv(ryear, rfs) ~ lp, 
                           gbsg5, 
                           reverse = TRUE,
                           timewt = "n/G2")

Uno_C_gbsg5_pgr <- concordance(Surv(ryear, rfs) ~ lp_pgr, 
                           gbsg5, 
                           reverse = TRUE,
                           timewt = "n/G2")
```
</details>
  
```{r, concordance, fig.align='center', warning=FALSE, eval=TRUE}
```


```{r, int_val_01, echo=FALSE, warning=FALSE}
source(here::here('Functions/internal_cv.R'))
# See legend of the argument in the Functions folder
```

Internal validation using bootstrap optimism-corrected cross-validation

<details>
  <summary>Click to expand code</summary>
```{r, int_val_calc_01, warning=FALSE, message=FALSE, echo=TRUE, eval=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

int_val <- bootstrap_cv(db = rott5, 
                        B = 10,
                        time = "ryear",
                        status = "rfs",
                        formula_model = "Surv(ryear, rfs) ~ csize + 
                        cnode + 
                        grade3",
                        formula_ipcw = "Surv(ryear, rfs) ~ 1",
                        pred.time = 4.99)

int_val_pgr <- bootstrap_cv(db = rott5, 
                        B = 10,
                        time = "ryear",
                        status = "rfs",
                        formula_model = "Surv(ryear, rfs) ~ csize + 
                        cnode + 
                        grade3 + pgr2 + pgr3",
                        formula_ipcw = "Surv(ryear, rfs) ~ 1",
                        pred.time = 4.99)
```
</details>
  
```{r, int_val_calc_01, fig.align='center', warning=FALSE, eval=TRUE}
```

NOTE: we use only B = 10 bootstrap samples. You can increase B although it is computationally more demanding. 

```{r, res_UnoC, echo=FALSE}
alpha <- .05
k <- 2
res_C <- matrix(c(
  harrell_C_rott5$concordance,
  harrell_C_rott5$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5$var) ,
  harrell_C_rott5$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5$var),
  
  int_val["Harrell C corrected"],
  NA,
  NA,
  
  harrell_C_rott5_pgr$concordance,
  harrell_C_rott5_pgr$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5_pgr$var) ,
  harrell_C_rott5_pgr$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5_pgr$var),
  
  int_val_pgr["Harrell C corrected"],
  NA,
  NA,
  
  harrell_C_gbsg5$concordance,
  harrell_C_gbsg5$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_gbsg5$var) ,
  harrell_C_gbsg5$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_gbsg5$var),
  
  harrell_C_gbsg5_pgr$concordance,
  harrell_C_gbsg5_pgr$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_gbsg5_pgr$var) ,
  harrell_C_gbsg5_pgr$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_gbsg5_pgr$var),
  
  Uno_C_rott5$concordance,
  harrell_C_rott5$concordance - 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5$var) ,
  harrell_C_rott5$concordance + 
    qnorm(1 - alpha / 2) * sqrt(harrell_C_rott5$var),
  
  int_val["Harrell C corrected"],
  NA,
  NA,
  
  Uno_C_rott5_pgr$concordance,
  Uno_C_rott5_pgr$concordance - 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_rott5_pgr$var) ,
  Uno_C_rott5_pgr$concordance + 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_rott5_pgr$var),
  
  int_val_pgr["Uno C corrected"],
  NA,
  NA,
  
 
  Uno_C_gbsg5$concordance,
  Uno_C_gbsg5$concordance - 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_gbsg5$var) ,
  Uno_C_gbsg5$concordance + 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_gbsg5$var),
  
  Uno_C_gbsg5_pgr$concordance,
  Uno_C_gbsg5_pgr$concordance - 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_gbsg5_pgr$var) ,
  Uno_C_gbsg5_pgr$concordance + 
    qnorm(1 - alpha / 2) * sqrt(Uno_C_gbsg5_pgr$var)
),
  
  nrow = 2,
  ncol = 18,
  byrow = T,
  
  dimnames = list(
    c("Harrell C", "Uno C"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"), 6)
  )
)

res_C <- round(res_C, 2) # Digits
kable(res_C) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, 
                     "Apparent" = 3, 
                     "Internal" = 3, 
                     "Apparent + PGR" = 3, 
                     "Internal + PGR" = 3, 
                     "External" = 3, 
                     "External + PGR" = 3))
```

Concordance varied between 0.67 and 0.68 in the apparent, internal and external validation using the basic and extended model.

<details>
  <summary>Click to expand code</summary>
```{r, AUC, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}

# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Time-dependent AUC (in Table 3 called Uno's TD AUC at 5 years) ###
# Uno's time-dependent Area Under the Curve
# Development
Uno_rott5 <-
  timeROC(
    T = rott5$ryear, 
    delta = rott5$rfs,
    marker = predict(efit1, newdata = rott5),
    cause = 1, 
    weighting = "marginal", 
    times = 4.99,
    iid = TRUE
  )

Uno_rott5_pgr <-
  timeROC(
    T = rott5$ryear, 
    delta = rott5$rfs,
    marker = predict(efit1_pgr, newdata = rott5),
    cause = 1, 
    weighting = "marginal", 
    times = 4.99,
    iid = TRUE
  )

# Validation
Uno_gbsg5 <-
  timeROC(
    T = gbsg5$ryear, 
    delta = gbsg5$rfs,
    marker = predict(efit1, newdata = gbsg5),
    cause = 1, 
    weighting = "marginal", 
    times = 4.99,
    iid = TRUE
  )

Uno_gbsg5_pgr <-
  timeROC(
    T = gbsg5$ryear, 
    delta = gbsg5$rfs,
    marker = predict(efit1_pgr, newdata = gbsg5),
    cause = 1, 
    weighting = "marginal", 
    times = 4.99,
    iid = TRUE
  )
# NOTE: if you have a lot of data n > 2000, standard error computation may be really long.
# In that case, please use bootstrap percentile to calculate confidence intervals.
```
</details>
  
```{r, AUC, fig.align='center', warning=FALSE, eval=TRUE}
```


```{r, res_AUC, echo=FALSE}
# Save results
alpha <- .05
k <- 2
res_AUC <- matrix(c(
 
  Uno_rott5$AUC["t=4.99"],
  Uno_rott5$AUC["t=4.99"] -
    qnorm(1 - alpha / 2) * Uno_rott5$inference$vect_sd_1["t=4.99"],
  Uno_rott5$AUC["t=4.99"] +
    qnorm(1 - alpha / 2) * Uno_rott5$inference$vect_sd_1["t=4.99"],
  
  int_val["AUC corrected"],
  NA,
  NA,
  
  Uno_rott5_pgr$AUC["t=4.99"],
  Uno_rott5_pgr$AUC["t=4.99"] -
    qnorm(1 - alpha / 2) * Uno_rott5_pgr$inference$vect_sd_1["t=4.99"],
  Uno_rott5_pgr$AUC["t=4.99"] +
    qnorm(1 - alpha / 2) * Uno_rott5_pgr$inference$vect_sd_1["t=4.99"],
  
  int_val_pgr["AUC corrected"],
  NA,
  NA,
  
  Uno_gbsg5$AUC["t=4.99"],
  Uno_gbsg5$AUC["t=4.99"] -
    qnorm(1 - alpha / 2) * Uno_gbsg5$inference$vect_sd_1["t=4.99"],
  Uno_gbsg5$AUC["t=4.99"] +
    qnorm(1 - alpha / 2) * Uno_gbsg5$inference$vect_sd_1["t=4.99"],
  
  Uno_gbsg5_pgr$AUC["t=4.99"],
  Uno_gbsg5_pgr$AUC["t=4.99"] -
    qnorm(1 - alpha / 2) * Uno_gbsg5_pgr$inference$vect_sd_1["t=4.99"],
  Uno_gbsg5_pgr$AUC["t=4.99"] +
    qnorm(1 - alpha / 2) * Uno_gbsg5_pgr$inference$vect_sd_1["t=4.99"]
),

  nrow = 1, 
  ncol = 18, 
  byrow = T,

  dimnames = list(
    c("Uno AUC"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"), 6)
)
)
res_AUC <- round(res_AUC, 2) # Digits
kable(res_AUC) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, 
                     "Apparent" = 3, 
                     "Internal" = 3, 
                     "Apparent + PGR" = 3, 
                     "Internal + PGR" = 3, 
                     "External" = 3, 
                     "External + PGR" = 3))

```

Time-dependent AUC at 5 years was between 0.71 and 0.72 in the apparent, internal and external validation for the basic and extended model.


### 2.2 Calibration
Calibration is the agreement between observed outcomes and predicted probabilities.
For example, in survival models, a predicted survival probability at a fixed time horizon _t_ of 80% is considered reliable if it can be expected that 80 out of 100 will survive among patients who received a predicted survival probability of 80%.
Calibration can be assessed at a fixed time point (e.g. at 5 years), and globally (considering the entire range of the data).
In addition, different level of calibration assessment can be estimated according to the level of information available in the data. 
When individual data of development and validation set are available, full assessment of calibration is possible. Calibration at fixed time point is possible when baseline hazard at fixed time point and coefficient are available. When only coefficients are available, assessment of calibration is limited.


Since different level of information may be available, different level of calibration can be estimated:  mean, weak, and moderate calibration.

+ Mean calibration can be estimated:  

  + at a fixed time point:
    + using the Observed and Expected ratio at time t. The             estimated Observed number of events is calculated by the         Kaplan-Meier curve, and the Expected number of events is         calculated as the average of the estimated predicted risk        at time _t_ (e.g. 5 years);  
    
  + time range:  
    + using the exponential of Poisson model intercept,                exp(a)=(O/E), with log of cumulative hazard as offset.  

+ Weak calibration can be estimated:

  + at a fixed time point  
    + using a new ‘secondary’ Cox model using all of the             validation  data with the PI from the development model as         the only covariate.  
    
  + time range:  
    + using slope as the coefficient of PI in Poisson model with       log of baseline cumulative hazard minus PI as offset.  


+ Moderate calibration can estimated:
  + at a fixed time point:  
    + using flexible calibration curve, complemented with ICI,         E50, E90.  

  + time range: 
    + Plot the observed / expected number of events over time          using Poisson model intercept, exp(a) with log of                cumulative hazard as offset across all time points to _t_ 
      (i.e. 5 years);  
    + Model relationship between predictions and observed risk in       external dataset using Poisson model. Plot cumulative            hazard from Poisson model versus cumulative hazard from          original Cox model. 

More detailed explanations are available in the [paper](https://www.medrxiv.org/content/10.1101/2022.03.17.22272411v1).
  
#### 2.2.1 Mean calibration
##### 2.2.1.1  Mean calibration - fixed time point
The mean calibration at fixed time point (e.g. at 5 years) can be estimated using the Observed and Expected ratio.
The observed is estimated using the complementary of the Kaplan-Meier curve at the fixed time point.
The expected is estimated using the average predicted risk of the event at the fixed time point.

<details>
  <summary>Click to expand code</summary>
```{r, mean_cal_t, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

##  Observed / Expected ratio at time t ------------
# Observed: 1-Kaplan Meier at time (t)
horizon <- 5
obj <- summary(survfit(Surv(ryear, rfs) ~ 1, 
                       data = gbsg5), 
               times = horizon)

pred <- 1 - predictSurvProb(efit1, 
                            newdata = gbsg5, 
                            times = horizon)

pred_pgr <- 1 - predictSurvProb(efit1_pgr, 
                            newdata = gbsg5, 
                            times = horizon)

OE <- (1 - obj$surv) / mean(pred)
OE_pgr <- (1 - obj$surv) / mean(pred_pgr)

```
</details>
  
```{r, mean_cal_t, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, res_OE, message=FALSE, warning=FALSE, echo=FALSE}
alpha <- 0.05
res_OE <- matrix(c(OE,
                   OE * exp(-qnorm(1 - alpha / 2) * sqrt(1 / obj$n.event)),
                   OE * exp(qnorm(1 - alpha / 2) * sqrt(1 / obj$n.event)),
                   
                   OE_pgr,
                   OE_pgr 
                   * exp(-qnorm(1 - alpha / 2) * sqrt(1 / obj$n.event)),
                   OE_pgr 
                   * exp(qnorm(1 - alpha / 2) * sqrt(1 / obj$n.event))
                   ),
                 
                 nrow = 1,
                 ncol = 6,
                 byrow = T,
                 dimnames = list(
                   c("OE ratio"),
                   rep(c("Estimate", "Lower .95", "Upper .95"),2))
)

res_OE <- round(res_OE, 2)
kable(res_OE) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, 
                     "External" = 3, 
                     "External + PGR" = 3))

```

Observed and Expected ratio is 1.04 (95% CI: 0.93 - 1.17) for the basic model and 1.02 (95% CI: 0.90 - 1.14) for the extended model.

##### 2.2.1.2  Mean calibration - time range assessment 
The mean calibration for the entire data is calculated using the approach proposed by [Crowson et al.](https://journals.sagepub.com/doi/abs/10.1177/0962280213497434) to estimate calibration-in-the-large and slope using Poisson regression.
It compares the observed to predicted event counts, amethod that is closely related to the standardized mortality ratio (SMR), common in epidemiology.
For mean calibration, we show the ratio which is the exponential of the intercept of the Poisson model (i.e. calibration-in-the-large).

<details>
  <summary>Click to expand code</summary>
```{r, mean_cal, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)

# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

### Calibration in the large -------------
p <- predict(efit1, newdata = gbsg5, type = "expected")

# With pgr
p_pgr <- predict(efit1_pgr, newdata = gbsg5, type = "expected")

# Calibration in the large
pfit1 <- glm(rfs ~ offset(log(p)), 
             family = poisson, 
             data = gbsg5,
             subset = (p > 0))

# With pgr
pfit1_pgr <- glm(rfs ~ offset(log(p_pgr)), 
             family = poisson, 
             data = gbsg5,
             subset = (p_pgr > 0))

int_summary <- summary(pfit1)$coefficients
int_summary_pgr <- summary(pfit1_pgr)$coefficients

```
</details>
  
```{r, mean_cal, fig.align='center', warning=FALSE, eval=TRUE}
```


```{r, res_Poisson_int, echo=FALSE,message=FALSE,warning=FALSE}
alpha <- 0.05
res_cal <- matrix(c(
    exp(int_summary[1]),
    exp(int_summary[1] - qnorm(1 - alpha / 2) * int_summary[2]),
    exp(int_summary[1] + qnorm(1 - alpha / 2) * int_summary[2]),
    
    exp(int_summary_pgr[1]),
    exp(int_summary_pgr[1] - qnorm(1 - alpha / 2) 
        * int_summary_pgr[2]),
    exp(int_summary_pgr[1] + qnorm(1 - alpha / 2) 
    * int_summary_pgr[2])
),

  nrow = 1, 
  ncol = 6, 
  byrow = T,
  dimnames = list(
  c("Calibration in-the-large"),
  rep(c("Estimate", "Lower .95 ", "Upper .95"), 2)
)
)

k <- 2
res_cal <- round(res_cal, k) # number of digits
kable(res_cal) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1,  
                     "External" = 3, 
                     "External + PGR" = 3))
```
The ratio is 1.06 (95% CI: 0.94 - 1.19) for the basic model and 1.02 (95% CI: 0.91 - 1.15) for the extended model. 


#### 2.2.2 Weak calibration 
##### 2.2.2.1 Weak calibration - fixed time point
For a fixed time point assessment of weak calibration, we can predict the outcome at 5 years for every patient but we need to determine the observed outcome at 5 years even for those who were censored before that time. One way to do this is to fit a new ‘secondary’ Cox
model using all of the validation data with the PI from the development model as the only covariate.

<details>
  <summary>Click to expand code</summary>
```{r, weak_cal_t, echo=TRUE, message=FALSE, warning=FALSE, eval=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

## Save elements needed to estimate concordance indexes
# Development data 
rott5$lp <- predict(efit1, newdata = rott5)
rott5$lp_pgr <- predict(efit1_pgr, newdata = rott5)

# Validation data 
gbsg5$lp <- predict(efit1, newdata = gbsg5)
gbsg5$lp_pgr <- predict(efit1_pgr, newdata = gbsg5)

# Basic model
gval <- coxph(Surv(ryear, rfs) ~ lp, data = gbsg5)

# Extended model with PGR
gval_prs <- coxph(Surv(ryear, rfs) ~ lp_pgr, data = gbsg5)
```
</details>
  
```{r, weak_cal_t, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, res_cal, message=FALSE, warning=FALSE, echo=FALSE}
alpha <- 0.05

res_cal <- matrix(c(gval$coef,
                    gval$coef - qnorm(1 - alpha / 2) * sqrt(gval$var),
                    gval$coef + qnorm(1 - alpha / 2) * sqrt(gval$var),
                    
                    gval_prs$coef,
                    gval_prs$coef -
                      qnorm(1 - alpha / 2) * sqrt(gval$var),
                    gval_prs$coef + 
                      qnorm(1 - alpha / 2) * sqrt(gval$var) 
                   ),
          
                 nrow = 1,
                 ncol = 6,
                 byrow = T,
                 dimnames = list(
                   c("Calibration slope"),
                   rep(c("Estimate", "Lower .95", "Upper .95"),2))
)

res_cal <- round(res_cal, 2)
kable(res_cal) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, 
                     "External" = 3, 
                     "External + PGR" = 3))

```
Calibration slope was 1.07 (95% CI: 0.82-1.32) and 1.20 (95% CI: 0.95-1.45) for the basic and extended model, respectively.

##### 2.2.2.2 Weak calibration - time range assessment
The calibration for the entire data is calculated using the approach proposed by [Crowson et al.](https://journals.sagepub.com/doi/abs/10.1177/0962280213497434) to estimate calibration-in-the-large and slope using Poisson regression.
For weak calibration, we show the calibration slope.

<details>
  <summary>Click to expand code</summary>
```{r, weak_cal, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

## Calibration in the large and calibration slope -------------
p <- predict(efit1, newdata = gbsg5, type = "expected")
lp5 <- predict(efit1, newdata = gbsg5, type = "lp")  #linear predictor

# With pgr
p_pgr <- predict(efit1_pgr, newdata = gbsg5, type = "expected")
lp5_pgr <- predict(efit1_pgr, newdata = gbsg5, type = "lp")  

# Calibration slope
logbase <- log(p) - lp5
logbase_pgr <- log(p_pgr) - lp5_pgr

pfit <- glm(rfs ~ lp5 + offset(logbase), 
             family = poisson, 
             data = gbsg5,
             subset= (p > 0))

pfit_pgr <- glm(rfs ~ lp5_pgr + offset(logbase_pgr), 
             family = poisson, 
             data = gbsg5,
             subset = (p_pgr > 0))


slope_summary <- summary(pfit)$coefficients
slope_summary_pgr <- summary(pfit_pgr)$coefficients

```
</details>
  
```{r, weak_cal, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, res_Poisson, echo=FALSE,message=FALSE,warning=FALSE}
alpha <- 0.05
res_cal <- matrix(c(

    slope_summary[2, 1],
    slope_summary[2, 1] - qnorm(1 - alpha / 2) * slope_summary[2, 2],
    slope_summary[2, 1] + qnorm(1 - alpha / 2) * slope_summary[2, 2],
    
    slope_summary_pgr[2, 1],
    slope_summary_pgr[2, 1] - 
      qnorm(1 - alpha / 2) * slope_summary_pgr[2, 2],
    slope_summary_pgr[2, 1] + 
      qnorm(1 - alpha / 2) * slope_summary_pgr[2, 2]
),

  nrow = 1, 
  ncol = 6, 
  byrow = T,
  dimnames = list(
  c("Calibration slope"),
  rep(c("Estimate", "Lower .95 ", "Upper .95"), 2)
)
)

k <- 2
res_cal <- round(res_cal, k) # number of digits
kable(res_cal) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1,  
                     "External" = 3, 
                     "External + PGR" = 3))
```

Calibration slope was 1.05 (95% CI: 0.80-1.30) and 1.16 (95% CI: 0.93-1.40) for the basic and extended model, respectively.


#### 2.2.3 Moderate calibration 
##### 2.2.3.1 Moderate calibration - fixed time point
Moderate calibration concerns whether among patients with the same predicted risk, the observed event rate equals the predicted risk. A smooth calibration curve of the observed event rates against the predicted risks is used for assessment of moderate calibration.
The relation between the outcome at a fixed time point and predictions can be visualised by plotting the predicted risk from another ‘secondary’ Cox model against the predicted risk from the development model. 
Moderate calibration at fixed time point can be assessed using flexible calibration curve, complemented with ICI, E50, E90 as suggested by [Austin et al](https://onlinelibrary.wiley.com/doi/full/10.1002/sim.8570).

Calibration curve: it is a graphical representation of calibration in-the-large and calibration. It shows:  
      
  + on the *x-axis* the predicted survival (or risk) probabilities at a fixed time horizon (e.g. at 5 years);  
      
  + on the *y-axis* the observed survival (or risk) probabilities at a fixed time horizon (e.g. at 5 years);  
      
  + The 45-degree line indicates the good overall calibration. 
      Points below the 45-degree line indicates that the model overestimate the observed risk. 
      If points are above the 45-degree line, the model underestimate the observed risk;
      The observed probabilities estimated by the Kaplan-Meier curves (in case of survival) or by the complementary of the Kaplan-Meier curves (in case of risk in absence of competing risks) are represented in terms of percentiles of the predicted survival (risk) probabilities.
      
+ Integrated Calibration Index (ICI):  it is the mean absolute difference between smoothed observed proportions and predicted
probabilities;  

+ E50 and E90 denote the median, the 90th percentile of the absolute difference between observed and predicted probabilities of the outcome at time *t*;  

<details>
  <summary>Click to expand code</summary>
```{r, cal_rcs_metrics, fig.align='center', echo=TRUE, eval=FALSE}
# Models  ---
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Calibration plot --------
# Basic model
gbsg5$pred <- 1 - predictSurvProb(efit1, 
                                  newdata = gbsg5, 
                                  times = 5)
gbsg5$pred.cll <- log(-log(1 - gbsg5$pred))

# Extended model
gbsg5$pred_pgr <- 1 - predictSurvProb(efit1_pgr, 
                                      newdata = gbsg5, 
                                      times = 5)
gbsg5$pred.cll_pgr <- log(-log(1 - gbsg5$pred_pgr))


# Estimate actual risk - basic model
vcal <- cph(Surv(ryear, rfs) ~ rcs(pred.cll, 3),
            x = T,
            y = T,
            surv = T,
            data = gbsg5
) 

# Estimate actual risk - extended model
vcal_pgr <- cph(Surv(ryear, rfs) ~ rcs(pred.cll_pgr, 3),
            x = T,
            y = T,
            surv = T,
            data = gbsg5
) 


dat_cal <- cbind.data.frame(
  "obs" = 1 - survest(vcal, 
                      times = 5, 
                      newdata = gbsg5)$surv,
  
  "lower" = 1 - survest(vcal, 
                        times = 5, 
                        newdata = gbsg5)$upper,
  
  "upper" = 1 - survest(vcal, 
                        times = 5, 
                        newdata = gbsg5)$lower,
  "pred" = gbsg5$pred,
  
  
   "obs_pgr" = 1 - survest(vcal_pgr, 
                      times = 5, 
                      newdata = gbsg5)$surv,
  
  "lower_pgr" = 1 - survest(vcal_pgr, 
                        times = 5, 
                        newdata = gbsg5)$upper,
  
  "upper_pgr" = 1 - survest(vcal_pgr, 
                        times = 5, 
                        newdata = gbsg5)$lower,
  
  "pred_pgr" = gbsg5$pred_pgr
  
)


# Flexible calibration curve - basic model
dat_cal <- dat_cal[order(dat_cal$pred), ]

par(xaxs = "i", yaxs = "i", las = 1)
plot(
  dat_cal$pred, 
  dat_cal$obs,
  type = "l", 
  lty = 1, 
  xlim = c(0, 1),
  ylim = c(0, 1), 
  lwd = 2,
  xlab = "Predicted probability",
  ylab = "Observed probability", bty = "n"
)
lines(dat_cal$pred, 
      dat_cal$lower, 
      type = "l", 
      lty = 2, 
      lwd = 2)
lines(dat_cal$pred, 
      dat_cal$upper,
      type = "l", 
      lty = 2, 
      lwd = 2)
abline(0, 1, lwd = 2, lty = 2, col = "red")
title("Basic model - validation data ")


# Flexible calibration curve - extended model
dat_cal <- dat_cal[order(dat_cal$pred_pgr), ]
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  dat_cal$pred_pgr, 
  dat_cal$obs_pgr,
  type = "l", 
  lty = 1, 
  xlim = c(0, 1),
  ylim = c(0, 1), 
  lwd = 2,
  xlab = "Predicted probability",
  ylab = "Observed probability", 
  bty = "n"
)
lines(dat_cal$pred_pgr, 
      dat_cal$lower_pgr, 
      type = "l", 
      lty = 2, 
      lwd = 2)
lines(dat_cal$pred_pgr, 
      dat_cal$upper_pgr,
      type = "l", 
      lty = 2, 
      lwd = 2)
abline(0, 1, lwd = 2, lty = 2, col = "red")
title("Extended model - validation data ")

# Numerical measures ---------------
# Basic model
absdiff_cph <- abs(dat_cal$pred - dat_cal$obs)

numsum_cph <- c(
  "ICI" = mean(absdiff_cph),
  setNames(quantile(absdiff_cph, c(0.5, 0.9)), c("E50", "E90"))
)

# Extended model ------
absdiff_cph_pgr <- abs(dat_cal$pred_pgr - dat_cal$obs_pgr)

numsum_cph_pgr <- c(
  "ICI" = mean(absdiff_cph_pgr),
  setNames(quantile(absdiff_cph_pgr, c(0.5, 0.9)), c("E50", "E90"))
)

## Bootstrap confidence intervals
## for the cal------
B <- 50 # Set B = 2000 although it takes more time
gbsg5_boot <- bootstraps(gbsg5, times = B)

# Bootstrap calibration measures
numsum_boot <- function(split) {
  
  pred <- 1 - predictSurvProb(efit1, 
                                  newdata = analysis(split), 
                                  times = 5)
  
  pred_pgr <- 1 - predictSurvProb(efit1_pgr, 
                                  newdata = analysis(split), 
                                  times = 5)
  
  pred.cll <- log(-log(1 - pred))
  pred.cll_pgr <- log(-log(1 - pred_pgr))
 
# Estimate actual risk - basic model
  vcal <- cph(Surv(ryear, rfs) ~ rcs(pred.cll, 3),
              x = T,
              y = T,
              surv = T,
              data = analysis(split)
) 
 
 vcal_pgr <- cph(Surv(ryear, rfs) ~ rcs(pred.cll_pgr, 3),
            x = T,
            y = T,
            surv = T,
            data = analysis(split)
) 
 
 # Save objects needed
db_cal_boot <- data.frame(
  "obs" = 1 - survest(vcal, 
                      times = 5, 
                      newdata = analysis(split))$surv,
                      
  "pred" = pred,
  
  "obs_pgr" = 1 - survest(vcal_pgr,
                          times = 5, 
                          newdata = analysis(split))$surv,
  
  "pred_pgr" = pred_pgr
)

absdiff_boot <- abs(db_cal_boot$obs - db_cal_boot$pred)
absdiff_boot_pgr <- abs(db_cal_boot$obs_pgr -
                          db_cal_boot$pred_pgr)

res_cal_boot <- data.frame(
  "ICI" = mean(absdiff_boot),
  "ICI_pgr" = mean(absdiff_boot_pgr),
  "E50" = quantile(absdiff_boot, probs = .5),
  "E50_pgr" = quantile(absdiff_boot_pgr, probs = .5),
  "E90" = quantile(absdiff_boot, probs = .9),
  "E90_pgr" = quantile(absdiff_boot_pgr, probs = .9)
)
    
}
numsum_b <- gbsg5_boot |>
  mutate(num_cal_boot = map(splits, numsum_boot),
         
         ICI = map_dbl(num_cal_boot, ~ .x$ICI),
         ICI_pgr = map_dbl(num_cal_boot, ~ .x$ICI_pgr),
         
         E50 = map_dbl(num_cal_boot, ~ .x$E50),
         E50_pgr = map_dbl(num_cal_boot, ~ .x$E50_pgr),
         
         E90 = map_dbl(num_cal_boot, ~ .x$E90),
         E90_pgr = map_dbl(num_cal_boot, ~ .x$E90_pgr)
         )

```
</details>
  
```{r, cal_rcs_metrics, fig.align='center', warning=FALSE, eval=TRUE}
```


```{r, res_numcal, message=FALSE,warning=FALSE, fig.align='center',include=TRUE, echo=FALSE}
alpha <- .05
res_numcal <- matrix(c(numsum_cph["ICI"],
                       quantile(numsum_b$ICI, probs = alpha / 2),
                       quantile(numsum_b$ICI, 
                                probs = 1 - alpha / 2),
                       
                       numsum_cph["E50"],
                       quantile(numsum_b$E50, probs = alpha / 2),
                       quantile(numsum_b$E50, 
                                probs = 1 - alpha / 2),
                       
                       numsum_cph["E90"],
                       quantile(numsum_b$E90, probs = alpha / 2),
                       quantile(numsum_b$E90, 
                                probs = 1 - alpha / 2),
                       
                       numsum_cph_pgr["ICI"],
                       quantile(numsum_b$ICI_pgr, 
                                probs = alpha / 2),
                       quantile(numsum_b$ICI_pgr, 
                                probs = 1 - alpha /2),
                       
                       numsum_cph_pgr["E50"],
                       quantile(numsum_b$E50_pgr, 
                                probs = alpha / 2),
                       quantile(numsum_b$E50_pgr, 
                                probs = 1 - alpha / 2),
                       
                       numsum_cph_pgr["E90"],
                       quantile(numsum_b$E90_pgr, 
                                probs = alpha / 2),
                       quantile(numsum_b$E90_pgr, 
                                probs = 1 - alpha / 2)),
                     nrow = 2,
                     ncol = 9,
                     byrow = T,
                     dimnames = list(
                       c("External data",
                         "External data + PGR"),
                       rep(
                         c("Estimate", "Lower.95", "Upper.95"),
                         3))
)

res_numcal <- round(res_numcal, 2)
kable(res_numcal) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1,  
                     "ICI" = 3, 
                     "E50" = 3,
                     "E90" = 3))
```

In the validation, ICI at 5 years was 0.03 and 0.02 for the basic and extended model, respectively.

##### 2.2.3.2 Moderate calibration - time range assessment
Moderate calibration over the time range can be assessed by plotting the observed / expected number of events over time     using Poisson model intercept, exp(a) with log of                cumulative hazard as offset across all time points to _t_ (here 5 years)

<details>
  <summary>Click to expand code</summary>
```{r, mod_cal, message=FALSE,warning=FALSE, fig.align='center', echo=TRUE, eval=FALSE}
# Models ------------
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Observed / Expected ratio across all time point to t = 5 years
# Time points
times_t <- seq(0.15, 5, by = .05)

# intercept across all time points - basic model 
int_t <- matrix(NA, 
                nrow = length(times_t),
                ncol = 4)
 
# intercept across all time points - extended model
int_t_pgr <- matrix(NA,
                    nrow = length(times_t),
                    ncol = 4)

# Estimate intercepts across time t using Poisson model
for(j in 1:length(times_t)) {
   # Administratively censoring data at time t
          temp <- survSplit(Surv(ryear, rfs) ~ ., 
                     data = gbsg5,
                     cut = times_t[j],
                     episode = "epoch") |>
            filter(epoch == 1)
          
  # Calculate cumulative hazard in every data
  # based on the basic model
           p <- predict(efit1,
                        newdata = temp,
                        type = "expected")
          
  # Calculate cumulative hazard in every data
  # based on the extended model
          p_pgr <- predict(efit1_pgr,
                           newdata = temp,
                           type = "expected")
          
  # Fit Poisson model to estimate the intercept
  # offsetting log cumulative hazard 
          
  # Basic model
          fit <- 
            glm(rfs ~ offset(log(p)), 
             family = poisson, 
             data = temp,
             subset = (p > 0 ))
          
  # Extended model
          fit_pgr <- 
            glm(rfs ~ offset(log(p_pgr)), 
             family = poisson, 
             data = temp,
             subset = (p > 0 ))
          
          # Save output
          int_t[j,] <- summary(fit)$coefficients
          int_t_pgr[j, ] <- summary(fit_pgr)$coefficients
          
}

# Colnames
colnames(int_t) <- colnames(summary(fit)$coefficients)
colnames(int_t_pgr) <- colnames(summary(fit_pgr)$coefficients)

# Results
res_OE_t <- cbind.data.frame(
  "time" = times_t,
  
  "OE" = exp(int_t[, "Estimate"]),
  
  "Lower.95" = exp(int_t[, "Estimate"] - 
                     qnorm(1 - alpha / 2) * 
                     int_t[, "Std. Error"]),
  
  "Upper.95" = exp(int_t[, "Estimate"] + 
                     qnorm(1 - alpha / 2) * 
                     int_t[, "Std. Error"]),
  
  "OE_pgr" = exp(int_t_pgr[, "Estimate"]),
  
  "Lower.95.pgr" = exp(int_t_pgr[, "Estimate"] - 
                     qnorm(1 - alpha / 2) * 
                     int_t_pgr[, "Std. Error"]),
  
  "Upper.95.pgr" = exp(int_t_pgr[, "Estimate"] + 
                     qnorm(1 - alpha / 2) * 
                     int_t_pgr[, "Std. Error"])
  
)

# Plot - basic model
par(xaxs = "i", yaxs = "i", las = 1)
oldpar <- par(mfrow = c(1, 2))
plot(res_OE_t$time,
     res_OE_t$OE,
     xlim = c(0, 5),
     ylim = c(0, 2),
     xlab = "Year of follow-up",
     ylab = 
       "Number of observed events / Number of expected events",
     cex.lab = .60,
     bty = "n",
     type = "l",
     col = "red",
     lwd = 2,
     main = " Basic model",
     cex.main = .70,
     cex.axis = .60)
lines(res_OE_t$time,
      res_OE_t$Lower.95,
      lwd = 2,
      lty = 2)
lines(res_OE_t$time,
      res_OE_t$Upper.95,
      lwd = 2,
      lty = 2)
abline (h = 1)

# Plot - extended model
par(xaxs = "i", yaxs = "i", las = 1)
plot(res_OE_t$time,
     res_OE_t$OE_pgr,
     xlim = c(0, 5),
     ylim = c(0, 2),
     xlab = "Year of follow-up",
     ylab = 
       "Number of observed events / Number of expected events",
     cex.lab = .60,
     bty = "n",
     type = "l",
     col = "red",
     lwd = 2,
     main = "Extended model",
     cex.main = .70,
     cex.axis = .60)
lines(res_OE_t$time,
      res_OE_t$Lower.95.pgr,
      lwd = 2,
      lty = 2)
lines(res_OE_t$time,
      res_OE_t$Upper.95.pgr,
      lwd = 2,
      lty = 2)
abline (h = 1)
```
</details>
  
```{r, mod_cal, fig.align='center', warning=FALSE, eval=TRUE}
```

We also plot the cumulative hazard from Poisson model versus cumulative hazard from original Cox model.

<details>
  <summary>Click to expand code</summary>
```{r, cumhaz_plot, message=FALSE,warning=FALSE, fig.align='center', echo=TRUE, eval=FALSE}

# Models ------------
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
               data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Cumulative hazard from Cox model - basic model
cumhaz_cox <- predict(efit1, 
                      newdata = gbsg5,
                      type = "expected")

lp <- predict(efit1,
              newdata = gbsg5,
              type = "lp")
logbase <- log(cumhaz_cox) - lp


# Poisson regression - basic model
fit_pois <- glm(rfs ~ lp + offset(logbase),
                family = poisson,
                data = gbsg5,
                subset = (cumhaz_cox > 0))

fit_pois <- glm(rfs ~ rcs(lp, 3)  +
                  offset(logbase),
                family = poisson,
                data = gbsg5,
                subset = (cumhaz_cox > 0))

cumhaz_pois <- predict(fit_pois,
                       type = "response", 
                       se.fit = TRUE)

dt_cumhaz <- cbind.data.frame(
  obs = cumhaz_cox[cumhaz_cox > 0],
  pred = cumhaz_pois$fit,
  se = cumhaz_pois$se.fit
)

dt_cumhaz <- dt_cumhaz[order(dt_cumhaz$pred), ]

alpha <- .05
par(xaxs = "i", yaxs = "i", las = 1)
plot(
  dt_cumhaz$obs, 
  dt_cumhaz$pred,
  type = "l", 
  lty = 1, 
  xlim = c(0, 2.5),
  ylim = c(0, 2.5), 
  lwd = 2,
  xlab = "Cumulative hazard from Cox model",
  ylab = "Cumulative hazard from Poisson model", 
  cex.lab = .70,
  cex.axis = .70,
  bty = "n",
  col = "red"
)
lines(dt_cumhaz$obs,
      dt_cumhaz$pred - qnorm(1 - alpha / 2) * dt_cumhaz$se,
      lwd = 2,
      lty = 2)
lines(dt_cumhaz$obs,
      dt_cumhaz$pred + qnorm(1 - alpha / 2) * dt_cumhaz$se,
      lwd = 2,
      lty = 2)
abline(a = 0, b = 1, lwd = 2)


```
</details>
  
```{r, cumhaz_plot, fig.align='center', warning=FALSE, eval=TRUE}
```

#### 2.2.4 Calibration when only coefficients of the model are available
When only coefficients of the development model is available and the baseline survival is not provided, only visual assessment of calibration is possible based on Kaplan-Meier curves between risk groups.

<details>
  <summary>Click to expand code</summary>
```{r, km, message=FALSE, warning=FALSE, echo=TRUE, eval=FALSE}

# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Development and validation dataset without pgr ---
rott5$lp <- predict(efit1, newdata = rott5,
                    reference = "zero")
rott5$group1 <- cut(rott5$lp, 
                    breaks = quantile(rott5$lp, 
                                      probs = seq(0, 1, 0.25)),
                    include.lowest = TRUE)

# Save breaks to be applied to external validation data
rott_breaks <- as.vector(quantile(rott5$lp, 
                                  probs = seq(0, 1, 0.25)))


gbsg5$lp <- predict(efit1, newdata = gbsg5, reference = "zero")
gbsg5$group1 <- cut(gbsg5$lp, 
                    breaks = c(
                      - 1000,
                      rott_breaks[2:4],
                      + 1000
                    ),
                    include.lowest = TRUE)


par(las = 1, xaxs = "i", yaxs = "i")
plot(survfit(Surv(ryear, rfs) ~ group1, data = gbsg5),
  bty = "n", 
  xlim = c(0, 5), 
  ylim = c(0, 1), 
  lwd = 2, 
  col = "black",
  lty = 2, 
  xlab = "Time (years)", 
  ylab = "Survival probability"
)
lines(survfit(Surv(ryear, rfs) ~ group1, data = rott5),
      lwd = 2)
legend("bottomleft",
       c("Development", "Validation"),
       lwd = 2, 
       lty = c(1, 2),
       bty = "n")
title("Basic model", adj = 0)

# Development and validation dataset with pgr ---
rott5$lp_pgr <- predict(efit1_pgr, newdata = rott5, 
                        reference = "zero")
rott5$group1_pgr <- cut(rott5$lp_pgr, 
                    breaks = quantile(rott5$lp_pgr, 
                                      probs = seq(0, 1, 0.25)),
                    include.lowest = TRUE)

# Save breaks to be applied to external validation data
rott_breaks_pgr <- as.vector(quantile(rott5$lp_pgr, 
                                  probs = seq(0, 1, 0.25)))

gbsg5$lp_pgr <- predict(efit1_pgr, newdata = gbsg5,
                        reference = "zero")
gbsg5$group1_pgr <- cut(gbsg5$lp_pgr, 
                    breaks = c(
                      - 1000,
                      rott_breaks_pgr[2:4],
                      + 1000
                    ) ,
                    include.lowest = TRUE)


par(las = 1, xaxs = "i", yaxs = "i")
plot(survfit(Surv(ryear, rfs) ~ group1_pgr, 
             data = gbsg5),
     bty = "n",
     xlim = c(0, 5), 
     ylim = c(0, 1), 
     lwd = 2, 
     col = "black",
     lty = 2, 
     xlab = "Time (years)", 
     ylab = "Survival probability"
)
lines(survfit(Surv(ryear, rfs) ~ group1_pgr, 
              data = rott5),
              lwd = 2)
legend("bottomleft",
       c("Development", "Validation"),
       lwd = 2, 
       lty = c(1, 2),
       bty = "n")
title("Extended model", adj = 0)
```
</details>
  
```{r, km, fig.align='center', warning=FALSE, eval=TRUE}
```

### 2.3 Overall performance measures
Two overall performance measures are proposed for prediction models with a survival outcome:

+ Brier score: it is the mean squared difference between observed event indicators and predicted risks at a fixed time point (e.g. at 5 years), lower is better;
  
+ Scaled Brier score, also known as Index of Prediction Accuracy (IPA): it improves interpretability by scaling the Brier Score. It is the decrease in Brier compared to a null model, expressed as a percentage, higher is better. 

<details>
  <summary>Click to expand code</summary>
```{r, overall, warning=FALSE, echo=TRUE, eval=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Development set (apparent Brier and IPA) without pgr
score_rott5 <-
  Score(list("cox_development" = efit1),
    formula = Surv(ryear, rfs) ~ 1, 
    data = rott5, 
    conf.int = TRUE, 
    times = 4.99,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )

# Validation set without pgr
score_gbsg5 <-
  Score(list("cox_validation" = efit1),
    formula = Surv(ryear, status) ~ 1, 
    data = gbsg5, 
    conf.int = TRUE,
    times = 4.99,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )

# Development set (apparent Brier and IPA) with pgr
score_rott5_pgr <-
  Score(list("cox_development" = efit1_pgr),
    formula = Surv(ryear, rfs) ~ 1, 
    data = rott5, 
    conf.int = TRUE, 
    times = 4.99,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )

# Validation set with pgr
score_gbsg5_pgr <-
  Score(list("cox_validation" = efit1_pgr),
    formula = Surv(ryear, rfs) ~ 1, 
    data = gbsg5, 
    conf.int = TRUE, 
    times = 4.99,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )

# Bootstrap for IPA -------------------------
B <- 10
set.seed(1234)
rott5_boot <- bootstraps(rott5, times = B)
gbsg5_boot <- bootstraps(gbsg5, times = B)

# IPA confidence intervals
# NOTE: computational time is too long to compute them when the number of bootstrap replications is high

# Bootstrap for IPA - basic model
score_boot <- function(split) {
  Score(list("model" = efit1),
    formula = Surv(ryear, rfs) ~ 1, 
    data = analysis(split), 
    conf.int = FALSE, 
    times = 4.99,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )$Brier$score[model == "model"][["IPA"]]
}

# Bootstrap for IPA - extended model with PGR
score_boot_pgr <- function(split) {
  Score(list("model" = efit1_pgr),
    formula = Surv(ryear, rfs) ~ 1, 
    data = analysis(split), 
    conf.int = FALSE, 
    times = 4.99,
    cens.model = "km", 
    metrics = "brier",
    summary = "ipa"
  )$Brier$score[model == "model"][["IPA"]]
}



rott5_boot <- rott5_boot |> mutate(
  IPA = map_dbl(splits, score_boot),
  IPA_pgr = map_dbl(splits, score_boot_pgr)
) # Development dataset

gbsg5_boot <- gbsg5_boot |> mutate(
  IPA = map_dbl(splits, score_boot),
  IPA_pgr = map_dbl(splits, score_boot_pgr)
) # Validation dataset

```
</details>
  
```{r, overall, fig.align='center', warning=FALSE, eval=TRUE}
```

```{r, int_val_02, echo=FALSE, warning=FALSE}
source(here::here('Functions/internal_cv.R'))
# See legend of the argument in the Functions folder
```

```{r, int_val_calc_02, echo=FALSE, warning=FALSE}
# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

int_val <- bootstrap_cv(db = rott5, 
                        B = 10,
                        time = "ryear",
                        status = "rfs",
                        formula_model = "Surv(ryear, rfs) ~ 
                        csize + 
                        cnode + 
                        grade3",
                        formula_ipcw = "Surv(ryear, rfs) ~ 1",
                        pred.time = 4.99)

int_val_pgr <- bootstrap_cv(db = rott5, 
                        B = 10,
                        time = "ryear",
                        status = "rfs",
                        formula_model = "Surv(ryear, rfs) ~ 
                        csize + 
                        cnode + 
                        grade3 + pgr2 + pgr3",
                        formula_ipcw = "Surv(ryear, rfs) ~ 1",
                        pred.time = 4.99)
```

```{r, overall_table, echo=FALSE}
# Table overall measures
alpha <- .05
k <- 2 # number of digits
res_ov <- matrix(unlist(c(
  score_rott5$Brier$score[model == "cox_development"][["Brier"]],
  score_rott5$Brier$score[model == "cox_development"][["lower"]],
  score_rott5$Brier$score[model == "cox_development"][["upper"]],
  
  int_val["Brier corrected"],
  NA,
  NA,
  
  score_rott5_pgr$Brier$score[model == "cox_development"][["Brier"]],
  score_rott5_pgr$Brier$score[model == "cox_development"][["lower"]],
  score_rott5_pgr$Brier$score[model == "cox_development"][["upper"]],
  
  int_val_pgr["Brier corrected"],
  NA,
  NA,
  
  score_gbsg5$Brier$score[model == "cox_validation"][["Brier"]],
  score_gbsg5$Brier$score[model == "cox_validation"][["lower"]],
  score_gbsg5$Brier$score[model == "cox_validation"][["upper"]],
  
  
  score_gbsg5_pgr$Brier$score[model == "cox_validation"][["Brier"]],
  score_gbsg5_pgr$Brier$score[model == "cox_validation"][["lower"]],
  score_gbsg5_pgr$Brier$score[model == "cox_validation"][["upper"]],
  
  score_rott5$Brier$score[model == "cox_development"][["IPA"]],
  quantile(rott5_boot$IPA, probs = alpha / 2),
  quantile(rott5_boot$IPA, probs = 1 - alpha / 2),
  
  int_val["IPA corrected"],
  NA,
  NA,
  
  score_rott5_pgr$Brier$score[model == "cox_development"][["IPA"]],
  quantile(rott5_boot$IPA_pgr, probs = alpha / 2),
  quantile(rott5_boot$IPA_pgr, probs = 1 - alpha / 2),
  
  int_val_pgr["IPA corrected"],
  NA,
  NA,
  
  score_gbsg5$Brier$score[model == "cox_validation"][["IPA"]],
  quantile(gbsg5_boot$IPA, probs = alpha / 2),
  quantile(gbsg5_boot$IPA, probs = 1 - alpha / 2),
  
  score_gbsg5_pgr$Brier$score[model == "cox_validation"][["IPA"]],
  quantile(gbsg5_boot$IPA_pgr, probs = alpha / 2),
  quantile(gbsg5_boot$IPA_pgr, probs = 1 - alpha / 2)
)),
  nrow = 2, 
  ncol = 18, 
  byrow = T, 
  dimnames = list(
    c("Brier", "IPA"),
    rep(c("Estimate", "Lower .95 ", "Upper .95"), 6)
)
)

res_ov <- round(res_ov, 2) # Digits
kable(res_ov) |>
  kable_styling("striped", position = "center") |>
  add_header_above(c(" " = 1, 
                     "Apparent" = 3, 
                     "Internal" = 3, 
                     "Apparent + PGR" = 3, 
                     "Internal + PGR" = 3, 
                     "External" = 3, 
                     "External + PGR" = 3))
```

As expected the overall performance measures were lower in the external validation. Including information about PGR slightly improved the overall performance.

## Goal 3 - Clinical utility
Discrimination and calibration measures are essential to assess the prediction performance but insufficient to evaluate the potential clinical utility of a risk prediction model for decision making. When new markers are available, clinical utility assessment evaluates whether the extended model helps to improve decision making.  
Clinical utility is measured by the net benefit that includes the number of true positives and the number of false positives. For example, in time-to-event models, the true positives reflect the benefit of being event free for a given time horizon using additional interventions such as additional treatments, personalized follow-up or additional surgeries. The false positives represent the harms of unnecessary interventions.   
Generally, in medicine, clinicians accepts to treat a certain number of patients for which interventions are unnecessary to be event free for a given time horizon. So, false negatives (the harm of not being event free for a given time horizon) are more important than false positives (the harm of unnecessary interventions). Thus, net benefit is the number of true positives classifications minus the false positives classifications weighted by a factor related to the harm of not preventing the event versus unnecessary interventions. The weighting is derived from the threshold probability to death (one minus survival probability) using a defined time horizon (for example 5 years since diagnosis). For example, a threshold of 10% implies that additional interventions for 10 patients of whom one would have experience the event in 5 years if untreated is acceptable (thus treating 9 unnecessary patients). This strategy is compared with the strategies of treat all and treat none patients. If overtreatment is harmful, a higher threshold should be used.  
The net benefit is calculated as:  
  
<img src="https://render.githubusercontent.com/render/math?math=%5Chuge%7B%5Cfrac%7BTP%7D%7Bn%7D-%5Cfrac%7BFP%7D%7Bn%7D*%5Cfrac%7Bp_t%7D%7B1-p_t%7D%7D">
  
*TP*=true positive patients   
*FP*=false positive patients  
*n*=number of patients and *p*<sub>t</sub> is the risk threshold.  

For survival data *TP* and *FP* is calculated as follows:   
<img src="https://render.githubusercontent.com/render/math?math=%5CLarge%7BTP%20%3D%20%5B1-S(t)%7C%20X%3D1%5D*P(X%3D1)*n%7D">
  
<img src="https://render.githubusercontent.com/render/math?math=%5CLarge%7BFP%20%3D%20%5BS(t)%7C%20X%3D1%5D*P(X%3D1)*n%7D">
  
  where  
*S(t)* survival at time *t*  
  *X=1* where the predicted probability at time *t* is  *p*<sub>t</sub>  
  
  And the the decision curve is calculated as follows:
  
1. Choose a time horizon (in this case 5 years);
2. Specify a risk threshold which reflects the ratio between harms and benefit of an additional intervention;
3. Calculate the number of true positive and false positive given the threshold specified in (2);
4. Calculate the net benefit of the survival model;
5. Plot net benefit on the *y-axis* against the risk threshold on the *x-axis*;
6. Repeat steps 2-4 for each model consideration;
7. Repeat steps 2-4 for the strategy of assuming all patients are treated;
8. Draw a straight line parallel to the *x-axis* at y=0 representing the net benefit associated with the strategy of assuming that all patients are not treated.

Given some thresholds, the model/strategy with higher net benefit represents the one that potentially improves  clinical decision making. However, poor discrimination and calibration lead to lower net benefit.

```{r, function_stdca, message=FALSE,warning=FALSE, fig.align='center',include=FALSE}
# Run the function to calculate the net benefit and the elements needed to develop decision curve analysis
source(here::here('Functions/stdca.R'))
```

<details>
  <summary>Click to expand code</summary>
```{r, dca, message=FALSE, warning=FALSE, fig.align='center', echo=TRUE, eval=FALSE}

# Models 
efit1 <- coxph(Surv(ryear, rfs) ~ csize + cnode + grade3,
  data = rott5, x = T, y = T
)
# Additional marker
efit1_pgr <- update(efit1, . ~ . + pgr2 + pgr3)

# Development data ----
# Predicted probability calculation
rott5$pred5 <- 1 - predictSurvProb(efit1, 
                                   newdata = rott5, 
                                   times = 5)

# Extended model with PGR
# Predicted probability calculation
rott5$pred5_pgr <- 1 - predictSurvProb(efit1_pgr, 
                                       newdata = rott5, 
                                       times = 5)

# Run decision curve analysis

# Development data
# Model without PGR
rott5 <- as.data.frame(rott5)
dca_rott5 <- stdca(
  data = rott5, 
  outcome = "rfs", 
  ttoutcome = "ryear",
  timepoint = 5, 
  predictors = "pred5", 
  xstop = 1.0,
  ymin = -0.01, 
  graph = FALSE
)

# Model with PGR
dca_rott5_pgr <- stdca(
  data = rott5, 
  outcome = "rfs", 
  ttoutcome = "ryear",
  timepoint = 5,
  predictors = "pred5_pgr",
  xstop = 1.0,
  ymin = -0.01, 
  graph = FALSE
)

# Decision curves plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(dca_rott5$net.benefit$threshold,
  dca_rott5$net.benefit$pred5,
  type = "l", 
  lwd = 2, 
  lty = 1,
  xlab = "Threshold probability in %", 
  ylab = "Net Benefit",
  xlim = c(0, 1),
  ylim = c(-0.10, 0.45), 
  bty = "n",
  cex.lab = 1.2, 
  cex.axis = 1
)
# legend('topright',c('Treat all','Treat none','Prediction model'),
#        lwd=c(2,2,2),lty=c(1,1,2),col=c('darkgray','black','black'),bty='n')
lines(dca_rott5$net.benefit$threshold, 
      dca_rott5$net.benefit$none, 
      type = "l", 
      lwd = 2, 
      lty = 4)
lines(dca_rott5$net.benefit$threshold, 
      dca_rott5$net.benefit$all, 
      type = "l",
      lwd = 2, 
      col = "darkgray")
lines(dca_rott5_pgr$net.benefit$threshold, 
      dca_rott5_pgr$net.benefit$pred5_pgr, 
      type = "l", 
      lwd = 2, 
      lty = 5)
legend("topright",
  c(
    "Treat All",
    "Original model",
    "Original model + PGR",
    "Treat None"
  ),
  lty = c(1, 1, 5, 4), 
  lwd = 2, 
  col = c("darkgray", "black", "black", "black"),
  bty = "n"
)
title("Development data", adj = 0, cex = 1.5)


# External data
# Validation data
# Predicted probability calculation
gbsg5$pred5 <- 1 - predictSurvProb(efit1, 
                                   newdata = gbsg5, 
                                   times = 5)

# Extended model with PGR
# Predicted probability calculation
gbsg5$pred5_pgr <- 1 - predictSurvProb(efit1_pgr, 
                                       newdata = gbsg5, 
                                       times = 5)

# Run decision curve analysis

# Development data
# Model without PGR
gbsg5 <- as.data.frame(gbsg5)
dca_gbsg5 <- stdca(
  data = gbsg5, 
  outcome = "rfs", 
  ttoutcome = "ryear",
  timepoint = 5, 
  predictors = "pred5", 
  xstop = 1.0,
  ymin = -0.01, 
  graph = FALSE
)
# Model with PGR
dca_gbsg5_pgr <- stdca(
  data = gbsg5, 
  outcome = "rfs", 
  ttoutcome = "ryear",
  timepoint = 5, 
  predictors = "pred5_pgr", 
  xstop = 1,
  ymin = -0.01, 
  graph = FALSE
)

# Decision curves plot
par(xaxs = "i", yaxs = "i", las = 1)
plot(dca_gbsg5$net.benefit$threshold,
  dca_gbsg5$net.benefit$pred5,
  type = "l", 
  lwd = 2, 
  lty = 1,
  xlab = "Threshold probability in %", 
  ylab = "Net Benefit",
  xlim = c(0, 1), 
  ylim = c(-0.10, 0.60), 
  bty = "n",
  cex.lab = 1.2, 
  cex.axis = 1
)
lines(dca_gbsg5$net.benefit$threshold, 
      dca_gbsg5$net.benefit$none, 
      type = "l", 
      lwd = 2, 
      lty = 4)
lines(dca_gbsg5$net.benefit$threshold, 
      dca_gbsg5$net.benefit$all, 
      type = "l", 
      lwd = 2, 
      col = "darkgray")
lines(dca_gbsg5_pgr$net.benefit$threshold,
      dca_gbsg5_pgr$net.benefit$pred5_pgr, 
      type = "l", 
      lwd = 2, 
      lty = 5)
legend("topright",
  c(
    "Treat All",
    "Original model",
    "Original model + PGR",
    "Treat None"
  ),
  lty = c(1, 1, 5, 4), lwd = 2, 
  col = c("darkgray", "black", "black", "black"),
  bty = "n"
)
title("External data", adj = 0, cex = 1.5)
```
</details>
  
```{r, dca, fig.align='center', warning=FALSE, eval=TRUE}
```


```{r, res_dca, echo=FALSE}
#read off at 23% threshold

# Development
# Validation
res_dca_dev <- round(cbind(
  dca_rott5$net.benefit[
    dca_rott5$net.benefit$threshold == 0.23, 
    c("threshold", "all", "pred5")],
  dca_rott5_pgr$net.benefit
  [dca_rott5_pgr$net.benefit$threshold == 0.23,
  c("pred5_pgr")]), 3)

rownames(res_dca_dev) <- NULL
colnames(res_dca_dev) <- c("Theshold", 
                       "Net benefit - Treat all",
                       "Net benefit - basic model",
                       "Net benefit - extended model with PGR")

# Validation
res_dca_val <- round(cbind(
  dca_gbsg5$net.benefit[
    dca_gbsg5$net.benefit$threshold == 0.23, 
    c("threshold", "all", "pred5")],
  dca_gbsg5_pgr$net.benefit
  [dca_gbsg5_pgr$net.benefit$threshold == 0.23,
  c("pred5_pgr")]), 3)

rownames(res_dca_val) <- NULL
colnames(res_dca_val) <- c("Theshold", 
                       "Net benefit - Treat all",
                       "Net benefit - basic model",
                       "Net benefit - extended model with PGR")

res_dca <- rbind(res_dca_dev, res_dca_val)
rownames(res_dca) <- c("Development data", "Validation data")
kable(res_dca) |>
  kable_styling("striped")
```

The potential net benefit at 23% threshold of the prediction model was 0.26, 0.27 for the basic and extended model in the development data, respectively.
This means that the model might identify approximately 26/27 patients out of 100 who may develop recurrence or may die within 5 years since diagnosis and thus adjuvant chemotherapy may help to reduce recurrence or mortality.
For validation data, the potential net benefit was 0.36 for the basic and extended model. 

Moreover, potential net benefit can be defined in terms of reduction of avoidable interventions (e.g adjuvant chemotherapy per 100 patients) by:
  
<img src="https://render.githubusercontent.com/render/math?math=%5Chuge%7B%5Cfrac%7BNB_%7Bmodel%7D%20-%20NB_%7Ball%7D%7D%7B(p_t%2F%20(1-p_t))%7D*100%7D%0A">  
  
  where *NB*<sub>model</sub> is the net benefit of the prediction model, *NB*<sub>all</sub> is the net benefit of the strategy treat all and $p_{t}$ is the risk threshold.

  
## Reproducibility ticket
```{r repro_ticket, echo=TRUE}
sessioninfo::session_info()
```
  
  
  
